---
title: "R for Data Science Summary Series"
author: "Fisher Ankney"
date: "January 17, 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data visualisation with ggplot2

This is the first of a five part series summarsing what I've learned while working through Hadley Wickham's R for Data Science: <http://r4ds.had.co.nz/>. 

I'd recommend Hadley's book as an excellent and most comprehensive introduction to R programming. Data visualisation is the first concept he introduces in the text. Data visualisation is incorporated throughout the entirety of R4DS, however it's most heavily covered in chapters 3, 7, and 28. The goal of this write up is to create a consise data visualisation reference guide based the concepts in these chapters. 

<br  />

#### Other Useful References    
R Graphics ggplot2 Cookbook - <http://www.cookbook-r.com/Graphs/>     
Top 50 Visualizations - <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>    
R Studio's ggplot2 Cheat Sheet - <https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf>    
Hadley's ggplot2 book (amazon) - <https://www.amazon.com/dp/331924275X/ref=cm_sw_su_dp>    

<br  />

#### Associated Libraries 
```{r message=FALSE}
library('tidyverse')
library('ggplot2')
library('RColorBrewer')
library('maps')
```
Be sure to install the packages on your computer using "install.packages('library_name')" before calling them with the library function. Tidyverse is an umbrella package that includes associated Hadley Wickham data science packages such as dplyr, tidyr, readr, etc. If you call for the tidyverse package, it is unnecessary to call again for the ggplot2 package. RColorBrewer provides easy access to an assortment of excellent color palettes. The maps package allows for simple geospatial plots of a select number of countries, including the continental United States. 

<br  />

#### Basic Template
```{r eval=FALSE}
ggplot( < enter data here > ) +            
geom_< enter geometry here >(              
  mapping = aes( < enter variables here > ),    
  stat = < enter statistical transformations here >,          
  position = < enter position adjustments here >        
) + 
coord_< enter coordinate system here >() +          
facet_< enter facet system here >()              
```
The basic idea here is that you can create and describe any possible graph with these seven parameters. I'll go through the most useful options for each of the necessary parameters, as well as the optional parameters that alter other aestetics like the legend and axes.

<br  /> 

#### ggplot
```{r}
ggplot(mpg)
```
     
As you can see, entering a dataset into the ggplot function isn't enough to produce a graph. ggplot() simply produces the base layer of the plot, you must specify geometry, mapping, aestetics, and so on to actually produce a plot. use 'ggplot() +' to link the base layer to these necissary specifications. 

<br  />

#### Geometry
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

the geom_( ) function does the majority of the grunt work in ggplot2. There are two necissary choices to be made with this function, choosing a geometry, and choosing the mappable variables. Additional commands are included within the geom( ) call, including aestetic specifications, statistical transformations, and position adjustments. 

##### Geometries    
* **geom_density( )** - 1 variable gaussian distribution
* **geom_point( )** - 2, 3, or 4 variable scatterplot 
* **geom_smooth( )** -  2 or 3 variable line plot
* **geom_bar(stat="identity")** -  2 or 3 variable bar plot
* **geom_hex( )** - 2 variable distribution
* **geom_tile( )** - 3 variable tile plot 
* **geom_map( )** - 2, 3, or 4 variable geospatial

##### Mappings 
* **aes( x = )** - x-variable
* **aes( ..., y = )** - y-variable
* **aes( ..., color / fill = )** typically a third or fourth variable 
* **aes( ..., size = )** typically a third or fourth variable

##### Aestetics 
* **size =** double (in millimeters)
* **linetype =** 0 (blank), 1 (solid), 1 (dashed), 3 (dotted), 4 (dot-dash), 5 (long-dash), 6 (two-dash)
* **weight =** (line thickness) double
* **color =** (outline) string
* **fill =** (inside) string
* **alpha =** (transparency) 0-1    

<br  />

#### Statistical Transformations    
Sometimes you want to visualize a transformed version of the original data. My most commonly used statistcal transformation is identity, use this transformation with geom_bar to visualize a variable instead of a count.   
    
* **geom_(aes(...), stat = "count")** - visualise the number of entires in a variable, n()    
* **geom_(aes(...), stat = "identity")** - visuaise the variable, not the count    
* **geom_(aes(...), stat = "unique")** - visualise only unique components of the variable    

<br  />

#### Position Adjustments
When geometric data occupies the same space, it is useful to use a position adjustment to better understand the visualisation.    
    
* **geom_point(aes(...), position = "jitter")** - adds a small amount of noise to better view overlapping points     
* **geom_bar(aes(...), position = "dodge")** - arranges bar elements side-by-side    
* **geom_bar(aes(...), position = "fill")** - arranges bar elements on top of each other, normalizing     height    

<br  />
   
#### Coordinate Systems
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  coord_flip()  
```
<br  />     
Sometimes it's useful to change the coordinate system in a particular plot. This will affect the axes and how the data is plotted. I find the most useful coordinate transformations to be coord_flip (typically with bar plots) and coord_trans (typically with exponential or logarithmic trends).    
    
* **coord_fixed(ratio, xlim, ylim)** - fixing the aspect ratio between x and y
* **coord_flip(xlim, wlim)** - flipping the axes 
* **coord_polar(theta, start, direction)** - coverting cartesian to polar coordinates
* **coord_trans(xtrans, ytrans, limx, limy)** - transform cartesian coordiantes, use with log / exponent functions    
* **coord_map(projection, orientation, xlim, ylim)** - mapproj package projections 
   
   
#### Facets

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
     
Dividing a plot into subplots can be helpful when you want to evaluate a dataset based on the values of one or more discrete variables. 

* **facet_grid(x ~ ., labeller, scales)** - facets the display into columns based on the given variable and placeholder "."    
* **facet_grad(. ~ x, labeller, scales)** - facets the display into rows based on the given variable and plcaeholder "."     
* **facet_grid(x ~ y, labeller, scales)** - facets the display into rows and columns based on the two given variables.    
* **facet_wrap(~ x, labeller, scales, nrow, ncol)** - wraps facets into a rectangular layout    

<br  /> 
     
#### Labels  
```{r}
ggplot(mpg) + 
  labs(
    title = "title", 
    subtitle = "subtitle",
    caption = "caption",
    xlab = "x-label", 
    ylab = "y-label"
  )
```
      

Labels are very important for effectively communication within your visualisation. Any label can contain mathematical formulas by using the quote( ) function. for more information on this, type ?plotmath into the R console.    

<br  />

#### Annotations     

It can be useful to label major compnents and interesting observations within your visualisation. To add annotations to your plot, call the geom_text( ) function and assigne the aes( label = ) to a variable of interest. I don't particularly care for annotating individual observations, as I find it clutters the graphic. 

* **geom_text(aes(label = "text here"), vjust = , hjust = )** - places a summary annotation within the graphic, according to vjust and hjust     
* **geom_hline(yintercept = , size = , color = , linetype)** - creates a horizontal reference line through the plot according to yintercept     
* **geom_vline(xintercept = , size =, color =, linetype =)** - creates a verticle reference line through the plot according to xintercept     
* **geom_rect(aes(xmin = , xmax = , ymin = , ymax = ))** - creates a rectangle to box in and highlight interesting data    
* **geom_segment(aes(x = , y = , xend = , yend = ,) arrow = )** - creates a line segment within the plot, it can be an arrow     

<br  />

#### Scales    
 
Scales are automatically set when creating a ggplot, but sometimes it's useful to alter the color scheme, legend, or axes to better communicate your visulisation. The naming sceme for scales is scale_ followed by the name of the aesthetic then _, and the name of the scale.

##### Aestetic Names
     
  * **\_x\_**    
  * **\_y\_**      
  * **\_alpha\_**      
  * **\_color\_**           
  * **\_fill\_**         
  * **\_linetype\_**         
  * **\_shape\_**          
  * **\_size\_**        

##### Scale Names     
      
  * **\_discrete** - maps discrete variables to visual values      
  * **\_continous** - maps continuous variables to visual values
  * **\_identity** - uses data values as visual values 
  * **\_manual** - values = c( ), maps discrete values to manually chose visual values 
  * **\_gradient** - creates a two color gradient, low - high
  * **\_gradient2** - creates a diverging two color gradient, positive - negative
  * **\_gray** -creates a gradient in gray
  * **\_brewer** - brewer( palette = " "), calls RColorBrewer palettes    
     
##### Axes     

It's pretty easy to alter axes in ggplot2. Here are some commented examples for common problems:    
```{r eval=FALSE}
scale_x_continuous(breaks = seq(0, 10, by = 1), labels = c(1:9, "ten"))  # define breaks by sequence, define labels of breaks

scale_x_discrete(breaks=c("ctrl", "trt1", "trt2"),            # defined tick marks
                 labels=c("Control", "Treat 1", "Treat 2"))   # new names
                 
expand_limits(x = c(0,8), y = c(0,8))                         # expand the limits of the graph to visualise specific values

scale_y_reverse()     # Reverse y-axis direction (zero on top)

element_blank()       # Hiding axis labels 

theme(axis.title.x = element_text(face="bold", color="red", size=20),  # change axis labels, font and style
      axis.text.x = element_text(angle=90, vjust=0.5, size=16))        # change tick mark label, font, style

theme(panel.grid.major=element_blank(),  # hiding major gridlines
      panel.grid.minor=element_blank())  # hiding minor gridlines

library(scales)   # changing the y-axis to log 10
scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))
```

##### Legends    
Legends can be a bit harder to wrangle. Here are examples to several common problems I have with legends.    

Controlling the position, or existance of the legend 
```{r eval=FALSE}
theme(legend.position = "left")
theme(legend.position = "top")
theme(legend.position = "bottom")
theme(legend.position = "right") # the default

theme(legend.position = "none") # no legend will be generated

geom_point(aes(...), show.legend=FALSE) # don't include this geom in the legend    
```
     

Hiding the legend's title
```{r eval=FALSE}
theme(legend.title = element_blank())
```
     
Changing the order of items in your legend. Declare the scale_ function based on your plot, use guides = for a simple reverse     
```{r eval=FALSE}
scale_fill_discrete(breaks=(c("item1", "item2", "item3")))

scale_fill_discrete(guide = guide_legend(reverse=TRUE))
```

Manually alter the color, breaks, labels, and names of your legend, again the scale_ function is based on your plot
```{r eval=FALSE}
scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                  name="Experimental\nCondition",
                  breaks=c("ctrl", "trt1", "trt2"),
                  labels=c("Control", "Treatment 1", "Treatment 2"))
```

##### Colors    

Some of my favorite preset colors include 

* thistle
* slateblue2
* orchid
* midnightblue
* deepskyblue
* lightslateblue    

The RColorBrewer package also has pleasing palettes
```{r}
RColorBrewer::display.brewer.all()
```
      
#### Themes      

Sometimes you don't want to go through the effort of creating your own theme, so ggplot2 has built in themes to change it up a little bit. There are many expansions created for ggplot2 themes, but heres four of the most popular base ggplot2 themes. 

* **theme_bw()** - A while background with gridlines
* **theme_grey()** - the default theme 
* **theme_classic()** - A white background with no gridlines
* **theme_minimal()** - mystery theme, check it out! 

#### Summary 
