---
title: "R for Data Science Summary Series I"
subtitle: "Data Visualisation with ggplot2"
author: "Fisher Ankney"
date: "January 19, 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the first of a five part series summarsing the core concepts of Hadley Wickham's R for Data Science: <http://r4ds.had.co.nz/>. 

I'd recommend Hadley's book as an excellent and most comprehensive introduction to R programming. Hadley immediately introduces data visualisation concepts in R4DS, giving students a fun and highly functional introduction to R. While the data visualisation is continously referenced throughout the text, it is heavily emphasized in chapters 3, 7, and 28. This write up is an attempt to synthesis this information in a valueable and concise manner. Of course, this isn't a conprehensive guide to ggplot2, rather it's a small collection of code snippets and ideas that I find solve 80% of my ggplot2 problems.     

For more useful resources, including advanced ggplot2 techniques, check out these links:       

R Graphics ggplot2 Cookbook - <http://www.cookbook-r.com/Graphs/>     
Top 50 Visualizations - <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>    
R Studio's ggplot2 Cheat Sheet - <https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf>    
Hadley's ggplot2 book (amazon) - <https://www.amazon.com/dp/331924275X/ref=cm_sw_su_dp>    

<br  />

### Associated Libraries 
```{r message=FALSE}
# install.packages('tidyverse', 'RColorBrewer', 'maps')

library('tidyverse')     # install ggplot2 and associated tidyverse packages
library('ggplot2')       # install only ggplot2
library('RColorBrewer')  # install for excellent color palettes
library('maps')          # install for geospatial visualisations
```
I recommend installing the entirity of the tidyverse whenever you're looking to work with and graph data. If you don't want access to the other useful tidyverse tools, it's fine to just call the ggplot2 package by itself. RColorBrewer is my favorite way to quickly create visualy pleasing color schemes, but there are many other packages out there to choose from. The maps package integrates with ggplot2 really well, but there are probably better geospatial packages out there.    

<br  />

### Graphing Template
```{r eval=FALSE}
ggplot( < enter data here > ) +            
geom_< enter geometry here >(              
  mapping = aes( < enter variables here > ),    
  stat = < enter statistical transformations here >,          
  position = < enter position adjustments here >        
) + 
coord_< enter coordinate system here >() +          
facet_< enter facet system here >()              
```
With these seven parameters you can create and describe any possible plot. This is the underlying tidyverse philosophy, giving a 'grammer to graphics' as it were. Now let's go through these parameters individually, and toss in a few additional aestetic parameters for good measure.    

<br  /> 

### Base Graph
```{r}
ggplot(mpg)
```
      
Just calling ggplot( ) with an assoicated dataset will create a canvas for your graphic, but it won't do much else. We really need to add a geometry to the ggplot recipe.    

<br  />

### Geometry
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```
     
Declaring a plot geometry is perhaps the most important part of a ggplot recipe; when you're doing exploratory data analysis I find  you typically don't have to go any further than calling the geom function and declaring it's mappable variables. I've listed the handful of geometries, mappings, and aestetics that I use most often. There are hundreds of geometries if you include the extensions that the ggplot community has created; if you can imagine it, there's probably a ggplot geometry that can do it.  

##### Geometries    
* **geom_density( )** - 1 variable gaussian distribution
* **geom_point( )** - 2, 3, or 4 variable scatterplot 
* **geom_smooth( )** -  2 or 3 variable line plot
* **geom_bar(stat="identity")** -  2 or 3 variable bar plot
* **geom_hex( )** - 2 variable distribution
* **geom_tile( )** - 3 variable tile plot 
* **geom_map( )** - 2, 3, or 4 variable geospatial

##### Mappings 
* **aes( x = )** - x-variable
* **aes( ..., y = )** - y-variable
* **aes( ..., color / fill = )** typically a third or fourth variable 
* **aes( ..., size = )** typically a third or fourth variable

##### Aestetics 
* **size =** size of a point, declared by a double
* **linetype =** 0 (blank), 1 (solid), 1 (dashed), 3 (dotted), 4 (dot-dash), 5 (long-dash), 6 (two-dash)
* **weight =** (line thickness) double
* **color =** (outline) string
* **fill =** (inside) string
* **alpha =** (transparency) 0-1    

<br  />

### Statistical Transformations    

Sometimes you want to visualize a transformed version of the original data. The statistical transformation I use most often is the identity trasformation with geom_bar. There are, however, several other useful transformations.       
    
* **geom_(aes(...), stat = "count")** - visualise the number of entires in a variable, n()    
* **geom_(aes(...), stat = "identity")** - visuaise the variable, not the count    
* **geom_(aes(...), stat = "unique")** - visualise only unique components of the variable    

<br  />

### Position Adjustments
When geometric data occupies the same space, it is useful to use a position adjustment to better understand the visualisation. I really only find adjusting the positioning of data useful when I'm working with bar charts and scatterplots.   
    
* **geom_point(aes(...), position = "jitter")** - adds a small amount of noise to better view overlapping points     
* **geom_bar(aes(...), position = "dodge")** - arranges bar elements side-by-side    
* **geom_bar(aes(...), position = "fill")** - arranges bar elements on top of each other, normalizing     height    

<br  />
   
### Coordinates
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  coord_flip()  
```
<br  />     
A cartesian coordinates are the go-to for 95% of my visualisations, but the coord_ function does more than call different coordinate systems. Often it's useful to fix the x-y ratio, or flip the axis of bar charts. When your working with exponential data, you may want to transform an axis to a logarithmic system. These useful functions and many more can be done by the coord_ function.    
    
* **coord_fixed(ratio, xlim, ylim)** - fixing the aspect ratio between x and y
* **coord_flip(xlim, wlim)** - flipping the axes 
* **coord_polar(theta, start, direction)** - coverting cartesian to polar coordinates
* **coord_trans(xtrans, ytrans, limx, limy)** - transform cartesian coordiantes, use with log functions    
* **coord_map(projection, orientation, xlim, ylim)** - mapproj package projections 
   
   
### Facets

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
     
I don't use facets that much; to be honest I find them pretty distracting. If you're a fan of contrasting minerature plots, you might find these faceting options useful.    

* **facet_grid(x ~ y, labeller, scales)** - facets the display into rows and columns based on the two given variables ("."" placeholder)    
* **facet_wrap(~ x, labeller, scales, nrow, ncol)** - wraps facets into a rectangular layout    

<br  /> 
     
### Labels  
```{r}
ggplot(mpg) + 
  labs(
    title = "title", 
    subtitle = "subtitle",
    caption = "caption",
    xlab = "x-label", 
    ylab = "y-label"
  )
```
      
Labels are helpful for properly communicating your visualisation. Including these five labels with your plots can help clarify your topic, variables, and sources.    

<br  />

### Annotations     
Annotations are another feature I don't use too often, but I'll include them anyways. I think annotating individual observations is a mistake, it clutters the graphic and takes away from the effectiveness of the visualisation. Packages like plotly and shiny do a better job of creating interactive graphics if you want individual labels. If you, however, insist on creating a ggplot with annotations, these might be useful.     
      
* **geom_text(aes(label = "text here"), vjust = , hjust = )** - places a summary annotation within the graphic, according to vjust and hjust     
* **geom_hline(yintercept = , size = , color = , linetype)** - creates a horizontal reference line through the plot according to yintercept     
* **geom_vline(xintercept = , size =, color =, linetype =)** - creates a verticle reference line through the plot according to xintercept     
* **geom_rect(aes(xmin = , xmax = , ymin = , ymax = ))** - creates a rectangle to box in and highlight interesting data    
* **geom_segment(aes(x = , y = , xend = , yend = ,) arrow = )** - creates a line segment within the plot, it can be an arrow     
     
<br  />
 
### Scales    
 
Scales are automatically set when you create a ggplot, but sometimes it's useful to alter the color scheme, legend, or axes to better communicate your visulisation. The naming sceme for these functions  is scale_ followed by the name of the aesthetic then another _, and then the name of the scale.

##### Aestetic Names
     
  * **\_x\_**    
  * **\_y\_**      
  * **\_alpha\_**      
  * **\_color\_**           
  * **\_fill\_**         
  * **\_linetype\_**         
  * **\_shape\_**          
  * **\_size\_**        

##### Scale Names     
      
  * **\_discrete** - maps discrete variables to visual values      
  * **\_continous** - maps continuous variables to visual values
  * **\_identity** - uses data values as visual values 
  * **\_manual** - values = c( ), maps discrete values to manually chose visual values 
  * **\_gradient** - creates a two color gradient, low - high
  * **\_gradient2** - creates a diverging two color gradient, positive - negative
  * **\_gray** -creates a gradient in gray
  * **\_brewer** - brewer( palette = " "), calls RColorBrewer palettes    
     
<br  />

#### Axes      
```{r eval=FALSE}
scale_x_continuous(breaks = seq(0, 10, by = 1), labels = c(1:9, "ten")) 
# define breaks by sequence, define labels of breaks

scale_x_discrete(breaks=c("ctrl", "trt1", "trt2"),            # defined tick marks
                 labels=c("Control", "Treat 1", "Treat 2"))   # new names
                 
expand_limits(x = c(0,8), y = c(0,8))                         # expand the limits of the graph to visualise specific values

scale_y_reverse()     # Reverse y-axis direction (zero on top)

element_blank()       # Hiding axis labels 

theme(axis.title.x = element_text(face="bold", color="red", size=20),  # change axis labels, font and style
      axis.text.x = element_text(angle=90, vjust=0.5, size=16))        # change tick mark label, font, style

theme(panel.grid.major=element_blank(),  # hiding major gridlines
      panel.grid.minor=element_blank())  # hiding minor gridlines

library(scales)   # changing the y-axis to log 10
scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))
```
Scales can be used to alter axes in ggplot2, it's a fairly straight forward process. These are some commented examples for fixing common behavioral issues you may have with your axes.   

<br  />     

#### Legends    
Legends can be a bit harder to wrangle than axes. Heere are examples to several common problems I have with legends.      
```{r eval=FALSE}

theme(legend.position = "left")    # legend to the left of plot
theme(legend.position = "top")     # legend above plot
theme(legend.position = "bottom")  # legend below plot
theme(legend.position = "right")   # the default
theme(legend.position = "none")    # no legend will be generated

geom_point(aes(...), show.legend=FALSE) # don't include this geom in the legend    

theme(legend.title = element_blank()) # hides the legends title

# changing the order of items in your legend, scale_ function is dependent upon the type of plot
scale_fill_discrete(breaks=(c("item1", "item2", "item3")))

# simply reverse the current item display in thelegend
scale_fill_discrete(guide = guide_legend(reverse=TRUE))

# manually alter color, break, label, and name of legend items.
scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                  name="Experimental\nCondition",
                  breaks=c("ctrl", "trt1", "trt2"),
                  labels=c("Control", "Treatment 1", "Treatment 2"))
```

<br  />

#### Colors    

Some of my favorite preset colors include 

* thistle
* slateblue2
* orchid
* midnightblue
* deepskyblue
* lightslateblue    

The RColorBrewer package also has pleasing color palettes
```{r eval=FALSE}
RColorBrewer::display.brewer.all()
```
  
<br  />   

#### Themes      

Sometimes you don't want to go through the effort of creating your own theme, so ggplot2 has built in themes built in. here are four basic themes that are excellent for exploratory analysis or creating a base to build off.   

* **theme_bw()** - A while background with gridlines
* **theme_grey()** - the default theme 
* **theme_classic()** - A white background with no gridlines
* **theme_minimal()** - mystery theme, check it out! 

<br  />    

### Summary    
This is just a fraction of what ggplot2 can do, but with this infromation I think you should be able to solve 80% of your graphing needs. Again, the references listed at the beginning of this write up are excellent, and you should take a look at them. If you want to learn more about R for Data Science written by Hadley Wickham, it's free to read at <http://r4ds.had.co.nz/>. If you're too lazy to read the entire book, check out the other posts in my R for Data Science Summary Series.    


Thanks for reading,     
- Fisher 