---
title: "R4DS Chapter 3 - Data Visualization"
author: "Fisher Ankney"
date: "2/27/2018"
output: html_document
---

### Introduction 

How to visualize your data using ggplot2. ggplot2 implements grammer of graphics, a coherent system for describing and building graphs. ggplot2 is one of the 8 core packages in the tidyverse, and can be easily loaded as such. 

```{r, message=F}
library('tidyverse')
```

When explicitly loading functions from packages you can use the formatting `package::function()`. 

<br> 

### First Steps

This is the dataset we'll be working to start off this chapter:
```{r}
mpg
```

<br> 

The `mpg` dataframe that comes pre-loaded with the ggplot2 package. To plot the `mpg` dataframe with `displ` on the x-axis and `hwy` on the right axis execute this code: 
```{r}
ggplot() + 
  geom_point(mpg, mapping = aes(x = displ, y = hwy))
```

<br> 

A few things - The `ggplot()` function must always be called, and first. The `+` is a pipe of sorts, allowing further description of the desired plot. Geom functions come after the `ggplot() +` and specify the geometry of the plot. In this case it's a scatter plot, `geom_point()`. the first arguement of geom_point defines the dataset used. This arguement can be placed into ggplot, but I prefer it in the geom. 

<br> 

### Aesthetic Mappings 

A third value can be added to a two dimensional scatterplot by mapping it to an aesthetic. Aesthetics are visual properties of the objects in your plots: shapes, sizes, colors, etc. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, 
                           y = hwy, 
                           color = class
                           )
             )
```

<br> 

Other non-variable aethetic changes can be made:
```{r}
ggplot() + 
  geom_point(data = mpg, 
             aes(x = displ,
                 y = hwy),
             color = "blue", 
             shape = 8, 
             size=5,
             alpha = 0.8
             )
```

<br> 

## Facets

Facets can also add additional variables. `facet_wrap()` allows you to facet your plot by a single variable. The first arguement is a formula, which is created with a `~` followed by a variables name. This variable should be discrete. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

<br> 

Double facet! Based on two variables, use the `facet_grid()` function. Faceting by drv and cyl, or drive (4wd, front, rear) and cylenders (4,5,6,8) creates the following graph. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid( drv ~ cyl)
```

<br> 

### Geometric Objects

Different plots use different geometries, ggplot2 calls these geoms. Scatterplots, bar charts, line graphs, maps, pie charts, they all have an assigned geom. To browse the geoms simply type `geom_` into the R terminal while the ggplot2 package is loaded. RStudio's auto complete feature will list every avaliable geom in the package. There are thousands of add-ons out in the wild web, if ggplot2 doesn't have what suits you. 

Different geoms take different `mapping` aesthetics. The linetype aesthetic wouldn't work too well with a scatteplot, but works great on `geom_smooth`.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

<br> 

Plots can even have multiple geoms all together:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

<br> 

However, this creates a repetative command, and is better done as:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

<br> 

### Statistical Transformations 

A new dataset, diamonds! 
```{r}
diamonds
```

<br> 

Statistical transformations are sometimes useful with box plots. Count, identity, and unique three types of statistical transformations that come in handy. Count is the default statistical transformation for geom_bar. 

Count 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

<br> 

Identity allows you to map the height of bars to the raw values of a y variable instead of the height of the bars being tied to a count. I create the y variable then impliment it within the `aes()` mapping. To make this work you must change the default statistical transformation.  

Identity 
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       5,
  "Good",       10,
  "Very Good",  12,
  "Premium",    7,
  "Ideal",      4
)

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

<br> 

### Position Adjustments 

You can color bar charts based on variables - 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

<br> 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

<br> 

Or perhaps more usefully: 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

<br> 

Position = "identity" places the bar chart sections directly on top off each other, overlapping them so they become more exact. 
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, color = clarity)) + 
  geom_bar(alpha = 0.2, position = "identity")
```

<br> 

Position = "fill" allows you to more accurately compare the proportional makeup of each bar:
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

<br> 


Poisiton = "dodge" places overlaps next to each other instead of stacking them:
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

<br> 

### Coordinate Systems

Sometimes it's useful to flip coordinate systems using the `coord_flip()` function. It makes for more space with these box plots: 
```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() + 
  coord_flip()
```

`coord_fixed`, `coord_polar`, `coord_trans` and `coord_quickmap` may also come in handy. `coord_quickmap()` set the aspect ratio corectly for mapping. 
```{r}
library("maps")

fr <- map_data("france")

ggplot(fr, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

<br> 

### Summary 

TEMPLATE - 
``` 
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
    mapping = aes(<MAPPINGS>),
    stat = <STAT>, 
    position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

Seven Parameters to describe any plot - <br> 

- Dataset <br> 
- Shape <br> 
- Mapping Aethetics <br> 
- Statistical Transformations <br> 
- Positioning <br> 
- Coordinate Systems <br> 
- Facets <br>

This concludes my notes on chapter 3. 
