---
title: "R4DS Chapter 4 - Workflow Basics"
author: "Fisher Ankney"
date: "2/27/2018"
output: html_document
---
<br>

### Coding Basics

Frustration is natural when you start programming in R, because it's a stickler for punctuation and even one character out of place makes it complain. A quick review of the basics - 

R the calculator: 
```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

<br> 

Assigning a variable 
```{r}
x <- 3 * 4
```

<br> 

Everything should be named in **snake_case**. CamelCase and naming.with.periods is bad; I don't like it. 

You can inspect an object by typing it's name:
```{r}
x
```

<br> 

### Functions 

Functions follow a specific formula: 
```{r, eval=FALSE}
function_name(arg1 = val1, arg2 = val2, ...)
```

<br> 

If you forget to finish a function properly, R will promt you with a `+` in the console. You can finish the function or hit escape and try again. 

<br> 

### Practice 
 
```{r, eval=FALSE}
library('tidyverse')
```

```{r, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r, eval=FALSE}
filter(mpg, cyl = 8)
```

```{r, eval=FALSE}
filter(diamonds, carat > 3)
```







