---
title: "R4DS Chapter 5 - Data Transformation"
author: "Fisher Ankney"
date: "2/28/2018"
output: html_document
---

<br> 

```{r, message=FALSE}
library('tidyverse')
library('nycflights13')
```

<br> 

### The Common Variable Types - <br> 

- Integer <br> 
- Double <br> 
- Character <br> 
- Date-Time <br> 
- Logical <br> 
- Factor <br> 
- Data <br> 
<br> 

### Learning the Dplyr Basics 
- `filter()` - pick observations by their values <br> 
- `arrange()` - reorder rows <br> 
- `select()` - pick variables by their values <br> 
- `mutate()` - create new variables from existing ones <br> 
- `summarise()` - collapse many values to a summary <br> 

Use all of these dplyr commands with the group_by() and you can organize anything!

<br> 

### Filter

```{r}
(dec_25 <- filter(flights, day == 25, month == 12))
```

<br> 

Logical Operators & (and) | (or) and ! (not). Find all of the flights in november OR december:
```{r}
filter(flights, month == 11 | month == 12)
```

<br> 

Note can't you do this as:
```{r}
filter(flights, month == 11 | 12)
``` 
This is because it finds months equal to 11 or 12, which is true, which is interpreted as '1', so It'll find all of the flights in january. 

<br> 

A shorthand for the above problem:
```
nov_dec <- filter(flights, month %in% c(11, 12))
```

<br> 

### Practice Time 

Find all flights that had an arrival delay of two or more hours
```{r}
filter(flights, arr_delay >= 120)
```

<br> 

Flew to Houston (IAH or HOU):
```{r}
filter(flights, dest == "IAH" | dest == "HOU")
```

<br> 

Were operated by United, American, or Delta
```{r}
filter(flights, carrier == "UA" | carrier == "AA" | carrier == "DL")
```

<br> 

Departed in summer (July, August, and September)
```{r}
filter(flights, month == 7 | month == 8 | month == 9)
```

<br> 

Arrived more than two hours late, but didn’t leave late
```{r}
filter(flights, arr_delay > 120 & dep_delay <= 0)
```

<br> 

Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
filter(flights, dep_delay >= 60 & arr_delay <= 30)
```

<br> 

Departed between midnight and 6am (inclusive)
```{r}
filter(flights, dep_time >= 0 & dep_time <= 600)
```

<br> 

Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify he code needed to answer the previous challenges?
```{r}
filter(flights, between(dep_time, 0, 600))
``` 

<br> 

How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

<br> 


### Arrange 

Arrange organizes the rows (default ascending) by the parameters you give it. 

```{r}
arrange(flights, desc(arr_delay))
```
 
 <br> 
 
 ### Practice Time
 
How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
arrange(flights, desc(is.na(dep_time)))
```

<br> 

Sort flights to find the most delayed flights.
```{r}
arrange(flights, desc(dep_delay))
```

<br> 

Find the flights that left earliest.
```{r}
arrange(flights, dep_time)
```

<br> 

Arrange flights to find the fastest flights
```{r}
arrange(flights, air_time)
```

<br> 

Which flights travelled the shortest distance?
```{r}
arrange(flights, desc(distance))
```

<br> 

### Select 

Often datasets have many variables, select allows you to quickly create a subset of these variables. 

```{r}     
flights %>% 
select(year, month, day)
```

<br> 

select(flights dataset, variables: year through day inclusively)
```{r}
select(flights, year:day)
```

<br> 

select(flights dataset, leave out varaibles year through day inclusively)
```{r}
select(flights, -(year:day))
```

<br>

Select helper functions - <br> 
- `starts_with("abc")`: matches names that begin with “abc”. <br> 
- `ends_with("xyz")`: matches names that end with “xyz”. <br> 
- `contains("ijk")`: matches names that contain “ijk”. <br> 
- `matches("(.)\\1")`: selects variables that match a regular expression. <br> 
- `num_range("x", 1:3)`: matches x1, x2 and x3. <br> 
- `everything()`: goes within select, keeps everything after defined observations <br> 

<br> 

### Rename

```{r}
rename(flights, tail_num = tailnum)
```

<br> 

### Mutate

Creates new columns that are the function of existing columns
```{r}
flights %>% 
  select(year:day, ends_with("delay"), distance, air_time) %>%
  mutate(gain = arr_delay - dep_delay, speed = distance / air_time * 60)
```

<br> 


If you want to destroy the old data and only keep the newly created data use transmute()
```{r}
flights %>% 
  transmute(
            gain = arr_delay - dep_delay,
            hours = air_time / 60,
            gain_per_hour = gain / hours
  )
```

<br> 

Modular arithmatic %/% (integer division) and remainder, %%,  can be useful: 
```{r}
transmute(flights, 
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100
)
```

<br> 

### Other Useful Functions

`lag()` - offset by -1
`lead()` - offset by +1 
`cumsum()`: running sum <br> 
`cumprod()`: cumulative product <br> 
`cummin()`: cumulative minimum <br> 
`cummax()`: cumulative maximum <br> 
`cummean()`: cumulative mean <br> 

<br> 

### Summarize

The last key verb is summarize(). It collapses a data frame to a single row:
```{r}
summarize(flights, delay = mean(dep_delay, na.rm = TRUE))
```

<br> 


used with `group_by()`:
```{r}
by_day <- group_by(flights, year, month, day)
summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

<br> 

### The Pipe

`%>%`, read as then, strings together dplyr and other functions:
```{r}
flights %>%   
  group_by(dest) %>%    
  summarise(         
    count = n(),       
    dist = mean(distance, na.rm = TRUE),      
    delay = mean(arr_delay, na.rm = TRUE)     
  ) %>%                
  filter(count > 20, dest != "HNL")        
```

<br> 

Measurements of spread include: <br> 
- `sd()`: mean squared deviation aka standard deviation <br> 
- `IQR()`: Inter Quartile Range <br> 
- `mad(x)`: Median absolute deviation <br> 

<br> 

Destinations with the most variability: 
```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))
```

<br> 

Which destinations have the most carriers?
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```

<br> 

Counts are super useful!
```{r}
not_cancelled %>% 
  count(dest)
```

<br> 

Weight variable (wt) will count (sum) the specified parameter:
```{r} 
not_cancelled %>% 
  count(tailnum, wt = distance)
```
 
<br> 
 
How many flights left before 5am? (these usually indicate delayed flights from the previous day)
```{r}
not_cancelled %>%
  group_by(year, month, day) %>%     
  summarise(n_early = sum(dep_time < 500))
```

<br> 

What proportion of flights are delayed by more than an hour?
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>%   
  summarise(hour_perc = mean(arr_delay > 60))  
```

<br> 

### Ungroup

Which carrier has the most delays? 
```{r}
num_delays <- not_cancelled %>%     
  group_by(carrier) %>%  
  summarise(num_delay = sum(arr_delay > 0)) %>%  
  arrange(desc(num_delay))
num_delays[1,]
```

<br> 

How fast was the fastest flight to denver? Really don't know why pipes arnt working on this one

```{r}
not_cancelled %>%
  filter(dest == "DEN") %>%
  arrange(desc(arr_delay)) %>%
  select(year:distance, -sched_arr_time) %>%
  mutate(speed = distance / (air_time / 60)) %>%
  arrange(desc(speed))
```

<br> 

Where do the flights go most often?
```{r}
num_flights <- flights %>%
  group_by(dest) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
num_flights
```

<br> 

How long, in total, were flights delayed on christmas day?

```{r}
flights %>%  
  filter(month == 12 & day == 25) %>%
  select(dep_delay) %>% 
  cumsum() %>%
  max(na.rm=TRUE)
```

<br> 

How many flights were cancelled in November?
```{r}
nov_cancel <- flights %>%
  filter(is.na(dep_delay) & month == 11)
count(nov_cancel)
```

<br> 

Come up with another approach that will give you the same output as 
not_cancelled %>% count(dest) without using count() how many planes went to each destination

```{r}
not_cancelled %>% 
  group_by(dest) %>%
  summarise(count = n())
```

<br> 

Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
flights %>% 
  group_by(year, month, day) %>%
  filter(is.na(dep_delay)) %>%
  count()
```

<br> 

Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
flights %>% group_by(carrier, dest) %>% summarise(n()))

```{r}
not_cancelled %>%
  group_by(carrier) %>%
  summarise(num_delay = sum(arr_delay > 0)) %>%
  arrange(desc(num_delay))
```

```{r}
flights %>% 
  group_by(carrier, dest) %>% 
  summarise(n())
```

<br> 

What does the sort argument to count() do. When might you use it?
order a vector or factor into ascending or descending order


Group mutate & filter
```{r}
flights %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```

<br> 

Find groups bigger than a threshold 
```{r}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
popular_dests
```

<br> 

Standardize to produce per group metrics
```{r}
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)
```

<br> 

This concludes data transformation chater 5 notes. 

