---
title: "R4DS Chapter 7 - Exploratory Data Analysis"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

<br> 

How to visualize and transformation to explore your data in a systematic way - 

The Exploratory Data Analysis Cycle <br>

 1. Generate questions about your data. <br>
 2. Search for answers by visualising, transforming, and modelling your data. <br>
 3. Use what you learn to refine your questions and/or generate new questions. <br>
 
 Often it pays to investigate: <br> 

 1. What type of variation occurs within my variables? <br> 
 2. What type of covariation occurs between my variables? <br> 
 
 <br> 
 
 ### Variation 
 
```{r, message=FALSE}
library('tidyverse')
```

<br> 

 To examine the distribution of a categorical variable, use a bar chart:
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```
<br>

The height of the bars displays how many observations occurred with each x value. Compute these values manually with dplyr::count():
 
```{r}
diamonds %>% 
  count(cut)
```

<br> 

To examine the distribution of a continuous variable, use a histogram:
```{r}
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.25)
```

<br>

Compute this by hand:
```{r}
diamonds %>% 
  count(cut_width(carat, 0.25))
```

<br> 

Zooming into diamonds of a lesser size:
```{r}
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)
```

<br> 

When overlaying mutliple plots, use geom_freqpoly, not geom_hist
```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

<br>

### Unusual Values

Outliers can be hard to see
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

<br> 

Adjust the scope of the graph: 
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

now we can see these three unusual values, lets grab them using dplyr
```{r}
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual
```

<br>

You can't have measurements of 0, and 31.8 ad 58.9 are also errors. REMOVE! You should always have a good reason to remove values, and disclose it. 

<br>

 ### Practice!
 
Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.
```{r}
ggplot(diamonds) + 
  geom_freqpoly(mapping = aes(x = x), binwidth = 0.5, color = "red") +
  geom_freqpoly(mapping = aes(x = y), binwidth = 0.5, color = "blue") +
  geom_freqpoly(mapping = aes(x = z), binwidth = 0.5, color = "green") +
    coord_cartesian(xlim = c(-10, 65), ylim = c(0, 50))
```
There are several other errors, x and y are width and depth, as 
 diamonds are symmetrical. 

<br>

Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 100) + 
  geom_vline(xintercept = 1500)
```
 almost no diamonds cost $1500??

<br> 

How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?
```{r}
carat_count <- diamonds %>%
  group_by(carat) %>%
  summarise(count = n()) %>%
  filter(carat == 0.99 | carat == 1)
carat_count
``` 
23 diamondss are ranked as 0.99 carats, 1558 diamonds are ranked as 1 carat lol

<br>

Covariation (relationship between two related variables)
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```
Hard to see how the price of diamonds vary with quality, because the overall count varies so much.

<br> 

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```
 density, which is the count standardised so that the area under each frequency polygon is one.
 
 <br>

Lets take a crack at it with geom_boxplot()
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```
 
 <br>
 
When you have ranked factors, sometimes it's useful to reorder them
```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
```

<br>

Reordered by median highway mpg
```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))
```

<br>

### Exercises


Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.
```{r}
library('nycflights13')

not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
  select(sched_dep_time, dep_time)

cancelled <- flights %>% 
  filter(is.na(dep_delay), is.na(arr_delay)) %>%
  select(sched_dep_time, dep_time)

cancel_plot<- ggplot() + 
  geom_freqpoly(mapping = aes(x = not_cancelled$sched_dep_time, y = ..density..)) + 
  geom_freqpoly(mapping = aes(x = cancelled$sched_dep_time, y = ..density..), color = "red")

cancel_plot
```

<br> 

 What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

```{r}
price_predict <- diamonds %>%
  mutate(size_sum = x + y + x) %>%
  arrange(desc(price))
head(price_predict)
tail(price_predict)
```
Size = Price

<br> 

### A Taiste of Models

```{r}
library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))

# once you take out the effect of carat, cut rules
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))

# consise code
diamonds %>% 
  count(cut, clarity) %>% 
  ggplot(aes(clarity, cut, fill = n)) + 
  geom_tile()

# learn more with the ggplot2 book on amazon
```