---
title: "R4DS Chapter 10 - Tibbles"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

<br> 

### Introduction

```{r, message=FALSE}
library('tidyverse')
```

<br> 

Tibbles are dataframes, but they tweak some older behavior to make life easier. Tibble is one of the eight core tidyverse packages. Coerce dataframes into tibbles using `as_tibble()`. Create new tibbles by hand, tibble automatically recycles inputs of length 1 as shown below:
```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

<br> 

`tibble()` never changes the type of inputs, the names of the variables, or creates row names. It's possible for non-syntactic names to go into a tibble using the backtick character:
```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)

tb
```

<br> 

### Tibbles vs Dataframes

Tibbles have a method of printing that shows only the first ten rows, and all of the columns that will fit on screen. Each column also reports it's type, a feature that's not included in dataframes. 


 view 10 tibble rows
options(tibble.print_max = 10, tibble.print_min = 10)
 but view all the tibble columns
options(tibble.width = Inf)

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

<br> 

Sometimes you need more output that the default display, here are a few options changes that can help. 

`options(tibble.print_max = n, tibble.print_min = m`: if more than `m` rows, print only `n` rows <br> 
`options(dplyr.print_min = Inf)`: always show all rows <br> 
`options(tibble.width = Inf)`: always print all columns <br> 
`View()`: RStudio's built in data viewer <br>

<br> 

### Subsetting 

Two tools: `$` and `[[`. If you want to pull out a signle variable by name, `$` works well. If you want to pul out a signle variable by name or position, `[[` is your guy. 

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df$x

df[[1]]
```

<br> 

If you want to use these extractions in a pipe, you need the special 
 placeholder `.`
```{r}
df %>% 
  .$x
```
 
<br> 

If you have some old function that won't play well with tibble, use this code to turn it into a dataframe: 
```{r, eval=FALSE}
class(as.data.frame())
```

### Practice! 

How can you tell if an object is a tibble?
```{r}
str(mtcars)
```

<br> 

Compare and contrast the following operations on a dataframe and a tibble. 

```{r}
df <- data.frame(abc = 1, xyz = "a")
tb <- tibble(abc = 1, xyz = "a")

df$x
tb$x

df[, "xyz"]
tb[, "xyz"]

df[, c("abc", "xyz")]
tb[, c("abc", "xyz")]
```

<br> 

Practice referring to non-syntactic names in the following dataframe
```{r}
annoying <- tibble(
  `1` = 1:10, 
  `2` = `1`  * 2 + rnorm(length(`1`))
)

annoying
```

Extract the variable called 1:
```{r}
annoying[[1]]
```

plot a scatterplot of 1 vs 2 
```{r}
ggplot(data = annoying) + 
geom_point(aes(x = `1`, y = `2`))
```

Createa  new column `3` which is `2` divided by `1`
```{r}
annoying$`3` = annoying$`2` / annoying$`1`

annoying
```


```{r}
annoying
names(annoying)<- c("one", "two", "three")
annoying
```

<br>

That's all for chapter 10. 
