---
title: "R4DS Chapter 11 - Data Import"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

<br> 

### Introduction

There are many functions to read in data, but we focus on `read_csv()`. First and most important arguement is the path to the file - 

`fake_data <- read_csv("data/data.csv")`

You can create comma delim files inline too: 
```{r}
read_csv("a,b,c
          1,2,3
          4,5,6")
```
<br>

Use comment = "" to drop all lines that start with specified symbol
```{r}
read_csv(" A comment I want to skip
         x,y,z
         1,2,3", comment = "")
```

<br>

You can use skip = n to skip the first n lines; 
```{r}
read_csv("The first line of metadata
         The second line of metadata
         x,y,z
         1,2,3", skip = 2)
```

<br> 

Use col_names = FALSE to tell it the csv has no heading: ```read_csv("1,2,3\n4,5,6", col_names = FALSE)``

You can alternatively add column names if you'd like: `read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))`
 
Denote missing values with na = `read_csv("a,b,c\n1,2,.", na = ".")`

Read tab seperated files and fixed width files with the same rules:
`read_tsv()` and `read_fwf()`.

<br> 

### Parsing a Vector  

`parse_logical()` and `parse_integer()` are useful functions in tidyr.There’s basically nothing that can go wrong with these parsers so I won’t describe them here further. `parse_double()` is a strict numeric parser, and `parse_number()` is a flexible numeric parser. These are more complicated than you  might expect because different parts of the world write numbers in different ways. `parse_character()` seems so simple that it shouldn’t be necessary.
 But one complication makes it quite important: character encodings. `parse_factor()` create factors, the data structure that R uses to represent categorical variables with fixed and known values. `parse_datetime()`, `parse_date()`, and `parse_time()` allow you to parse various date & time specifications. These are the most complicated because there are so many different ways of writing dates.

Parsing Numbers
```{r}
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))

parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")

parse_number("$123,456,789")
parse_number("123.456.789", locale = locale(grouping_mark = "."))
parse_number("123'456'789", locale = locale(grouping_mark = "'"))
```

<br> 


Parsing Date-Times
```{r}
parse_datetime("2010-10-01T2010")
 If time is omitted, it will be set to midnight
parse_datetime("20101010")
> [1] "2010-10-10 UTC"

parse_date("01/02/15", "%m/%d/%y")
> [1] "2015-01-02"
parse_date("01/02/15", "%d/%m/%y")
> [1] "2015-02-01"
parse_date("01/02/15", "%y/%m/%d")
> [1] "2001-02-15"
```

Really just come here if you're having trouble reading in a file...

Export a csv to excel spreadsheet - `write_excel_csv()`

Write out data as a csv - `write_csv(data_frame, "saved_name.csv")`, 
`write_csv(challenge, "challenge.csv")`

`write_rds()` and `read_rds()` are uniform wrappers around the base
functions `readRDS()` and `saveRDS()`. These store data in R’s custom binary format called RDS. 


### Other types of data 

To get other types of data into R, we recommend starting with the tidyverse packages listed below. They’re certainly not perfect, but they are a good place to start. For rectangular data:
  
- haven: reads SPSS, Stata, and SAS files. <br> 
- readxl: reads excel files (both .xls and .xlsx) <br> 
- DBI: along with a database specific backend (e.g. RMySQL, RSQLite, RPostgreSQL etc) allows you to run SQL queries against a database and return a data frame. <br> 


For hierarchical data: use jsonlite (by Jeroen Ooms) for json, and xml2 for XML. Jenny Bryan has some excellent worked examples at https://jennybc.github.io/purrr-tutorial/.

That concludes chapter 11 notes. 

  

