---
title: "R4DS Chapter 12 - Tidy Data"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

### What is Tidy Data

THREE GOLDEN RULES OF A TIDY DATASET
1. Each variable must have its own column. <br>
2. Each observation must have its own row. <br>
3. Each value must have its own cell. <br>

You'll satisfy these rules by remembering to put each dataset in a tibble and put each variable in a column. 

The two most important functions in tidyr are `gather()` and `spread()`.

### Gather 
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

<br> 

The first entries are the set of columns that are screwed up, in this case 1999 and 2000 represent values not variables next we enter the key, which is the name of the variable whose values form column names i.e. the key is year (1999, 2000) finally we clean up the variable whose values are spread across the dataset, so in the problem columns (1999, 2000), there are a number of values that now need a new home. these values are named "cases" by the code above. 

Noow trying it for the second dataset
```{r}
table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
```

Join the two tables togather!
```{r}
tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
left_join(tidy4a, tidy4b)
```

### Spread
The opposite of gather, usable when the observations are on multiple roles, like this: 
```{r}
table2

spread(table2, key = type, value = count)
```

<br> 

With spread, we only need two parameters, key and value The column that contains variable names, the key column. Here, it’s type. The column that contains values forms multiple variables, the value column. Here it’s count.

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

<br> 

They cannot be symmetrical because you're defining the column headers to be the years (2015, 2016) and the values to be the return numbers. then youre trying to gather up column headers year and return, when they no longer exist. 

Tidy this tibble!
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12 
) 

preg %>%   
  gather("male", "female", key = "sex", value = pregnants) %>% 
  spread(key = "pregnant", value = "pregnants") %>%
  rename("not_pregnant" = "no", "pregnant" = "yes") %>%
  select(sex, pregnant, not_pregnant)
```

<br> 

### Seperate

`separate()` pulls apart one column into multiple columns by splitting wherever a separator character appears
```{r}
table3
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

<br> 


By default the spererator is a non-alphanumeric chr. but you can set it: 
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```

<br>

### Unite

`unite()` is the inverse of `separate()`: it combines multiple columns into a single column
```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```


### Tidy Arguements

Dropping extra values
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra="drop")
```

<br> 


Fills in the missing column with an NA placeholder (right side, column 3)
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill ="right")
```


explicit and implicit missing data
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```
return for 4th Q is NA (explicit), entire 2016 Q1 missing (implicit)

```{r}
stocks %>% 
  spread(year, return)

stocks %>% 
  complete(year, qtr)
```
`complete()` takes a set of columns, and finds all unique combinations. It then ensures the original dataset contains all those values, filling in explicit NAs where necessary.

Sometimes you know what should be in those missing values because of the previous entries. `fill()` will cary forward the last value of that column. 
```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment %>% 
  fill(person)
```

<br> 

Tidying in action 
```{r}
tidy_who <- who %>%   
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm=TRUE) %>% 
  mutate(key = stringr::str_replace(key,"newrel", "new_rel")) %>%  
  separate(key, c("new", "type", "sexage"), sep = "_")   

tidy_who %>% count(new) 

tidy_who <- select(tidy_who, -new, -iso2, -iso3) 

tidy_who %>%
  separate(sexage, c("sex", "age"), sep = 1) 
```

<br> 

 That concludes chapter 12 notes. 



