---
title: "R4DS"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

<br> 

```{r}
library('tidyverse')
library('nycflights13')
```

#### Four relational datasets <br> 
- airlines <br> 
- airports <br> 
- planes <br> 
- weather <br> 

Combined to create `flights`

<br> 

### Exercise 1

1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

Airports > name, lat, long, alt, Planes > tailnum


2. I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

Weather only relates to the nyc airports, that should be signified on the diagram

3. Weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?
 
It would be relational to arrival delay, airtime, dest, distance. 

4. We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?

Use `filter()` to pick out the special days, n() < threshold passenger number

<br> 

### Keys

The variables used to connect each pair of tables

primary key: `planes$tailnum` - uniquely identifies an observation in it's own table

foreign key: `flights$tailnum` - uniquely identifies an observation in another table

It's good practice to very primary keys do indeed uniquely identify each observations
```{r}
planes %>% 
  count(tailnum) %>%
  filter(n > 1)
``` 
No variables are double counted, good!

<br>

```{r}
weather %>% 
  count(year, month, day, hour, origin) %>%
  filter(n > 1)
```
No variables are double counted, good!

<br> 

Sometimes a table has no primary keys: 
```{r}
flights %>% 
  count(year, month, day, tailnum) %>% 
  filter(n > 1) 
```
Planes are used multiple times a day!

<br> 

### Mutating Joins

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
View(flights2)
```

<br> 

Adding the full names of the airlines to flights2
```{r}
flights2 %>% 
  select(-origin, -dest) %>% 
  left_join(airlines, by="carrier")

# called a mutating join because you could have done it with
# dplyr mutate: 

flights2 %>%
  select(-origin, -dest) %>% 
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

<br>

inner join: 
the simplest type of join, matching pairs of observations
whenever the keys are equal.

UNMATCHED ROWS ARE NOT INCLUDED IN THE RESULT

looks to me like inner and right join are stupid, you should be using `left_join()` or `full_join()`. `left_join()` preserves original dataset, only suplimenting with new "right-side" dataset. `full_join()` is if youre combining two equally important datsets and you need to key track of missing observations from both `left_join(dataset, by=NULL)`  natural join, uses all variables that appear in both tables.


### Practice 

Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays.
```{r}
flights2 <- flights %>%
  group_by(dest) %>%
  summarise(avg_delay = mean(arr_delay, na.rm=TRUE))

airports2 <- airports %>%  
  left_join(flights2, c("faa" = "dest")) %>% 
  filter(avg_delay != is.na(TRUE))
airports2 

avg_delay_by_dest <- airports2 %>%
  ggplot(aes(lon, lat)) +
  borders("state") +
  geom_point(aes(
                 color=avg_delay
                 )
            ) +
  scale_color_distiller(palette="RdYlGn") + 
  coord_quickmap()
avg_delay_by_dest
```

<br>


2. Add the location of the origin and destination (i.e. the lat and lon) to flights.
```{r}
flights_2 <- flights %>% 
  select(year:day, origin:dest)
airports_2 <- airports %>%
  select(faa, lat, lon)

origin_dest_location <- flights_2 %>%
  left_join(airports_2, c("dest" = "faa")) %>% 
  rename(dest_lat = lat, 
         dest_lon = lon
            ) %>% 
  left_join(airports_2, c("origin" = "faa")) %>%
  rename(origin_lat = lat, 
         origin_lon = lon) %>%
  select(year:origin, origin_lat, origin_lon, dest:dest_lon)
View(origin_dest_location)
```

<br> 


 3. Is there a relationship between the age of a plane and its delays?
```{r}
age_delay <- flights %>%
  select(-year) %>%
  left_join(planes, key = tailnum) %>% 
  select(tailnum, arr_delay, dep_delay, year) %>%
  filter(dep_delay < 250)

age_delay <- head(age_delay, 500)

ggplot(age_delay) + 
  geom_point(aes(x =year,
                 y=arr_delay
                 ))
```
 not that I can see. could do a cor.test() to make sure though

<br> 

### Filtering Joins

`semi_join(x,y)`   keeps all observations in x that have a match in y
`anti_join(x,y)`   drops all observations in x that have a match in y


Find top 10 destinations
```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest
```

<br> 

Find flights that went to these top 10.
```{r}
flights %>% 
  filter(dest %in% top_dest$dest)
```

<br> 

But thats difficult, use semi_join instead, basically drops non-top destinations.
```{r}
flights %>% 
  semi_join(top_dest)
```

<br> 

That's chapter 13. 
  