---
title: "R4DS Chapter 14 - Strings"
author: "Fisher Ankney"
date: "3/3/2018"
output: html_document
---

```{r, message=FALSE}
library('tidyverse')
library('stringr')
```


double_quote <- "\""  or '"'
single_quote <- '\''  or "'"
new_line <- "\n"
tab <- "\t"

<br> 

All stringr functions start with str_ so you can easily browse them: <br>
str_length("Hello, world!")                 counts number of characters <br>
str_c("x", "y", "z", sep = " and then ")    combines strings with seperator <br> 

<br> 

Replacing NA <br> 
`x <- c("abc", NA)` <br> 
`str_c("|-", x, "-|")` > "|-abc-|" NA <br> 
`str_c("|-", str_replace_na(x), "-|")` > "|-abc-|" "|-NA-|" <br> 
 
<br>

Strings are vectorized 
`str_c("prefix-", c("a", "b", "c"), "-suffix")`

<br>


Strings of length 0 are silently dropped
```{r}
name <- "Fisher"
time_of_day <- "evening"
birthday <- FALSE
 often useful with the if
str_c(
  "Good ", time_of_day, " ", name,
  if (birthday) " and HAPPY BIRTHDAY",
  "."
)
```

<br> 

Collapsing strings
`str_c(c("x", "y", "z"), collapse = ", ")`

<br> 

String subsetting 
`x <- c("Apple", "Banana", "Pear")`
`str_sub(x, 1, 3)`          show characters 1,2,3 in x vector
`str_sub(x, -3, -1)`      show characters 3rd from last, 2nd, last

<br> 

Matching expressions
`x <- c("apple", "banana", "pear")`
`str_view(x,"an")`
 `.` is a wildcard
`str_view(x, ".a.")`
 \\. to actually search for .

<br> 

Anchors
`^` to match the start of the string.
`$` to match the end of the string.
`str_view(x, "a$")`

Other special tools: <br> 
- \d: matches any digit. <br> 
- \s: matches any whitespace (e.g. space, tab, newline). <br> 
- [abc]: matches a, b, or c. <br> 
- [^abc]: matches anything except a, b, or c. <br> 
- `&` and `|` work here too, `str_view(c("grey", "gray"), "gr(e|a)y")`

<br> 

Detecting matches: `x <- c("apple", "banana", "pear")`, `str_detect(x, "e")`

How many common words start with t?
`sum(str_detect(words, "^t"))`

What proportion of common words end with a vowel?
`mean(str_detect(words, "[aeiou]$"))`

How many matches are in a string?
`x <- c("apple", "banana", "pear")`
`str_count(x, "a")`

<br> 

Really just use this chapter as reference if you run into strings. 




