---
title: "R4DS Chapter 18 - Pipes"
author: "Fisher Ankney"
date: "2/26/2018"
output: html_document
---
<br>

Pipes are a powerful tool for clear and decisive coding. 

This is a pipe: `>%>` 

This is where pipes come from: `library('magrittr')`

<br> 

### Many Ways to Foo Foo 

There are several ways to *program* this childrens poem:

```{r, eval=FALSE}
Little bunny Foo Foo
Went Hoping through the forest
scooping up the field mice
and bopping them on the head
```

<br> 

#### Save each step as a new object
```{r, eval=FALSE}
foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)
```

Cluttered code and a lot of typing; careful incrementation of each line is a hassle. 

<br>

#### Overwrite the original many times
```{r, eval=FALSE}
foo_foo <- hop(foo_foo, through = forest)
foo_foo <- scoop(foo_foo, up = field_mice)
foo_foo <- bop(foo_foo, on = head)
```

A bit less typing and thinking, but debugging is painful and the repitition obscures what is changing in each line. 

<br> 

#### Compose Functions
```{r, eval=FALSE}
bop(
  scoop(
    hop(foo_foo, through = forest),
    up = field_mice
  ), 
  on = head
)
```

Oh god, inside-out, right-to-left, this nest of functions is nearly impossible to comprehend. This is why people hate programming. 

<br>

#### Use the Pipe
```{r, eval=FALSE}
foo_foo %>% 
  hop(through = forest) %>%
  scoop(up = field_mouse) %>%
  bop(on = head) 
```

Simple, elegant, easy to read and create. 

<br> 

### When Not To Use The Pipe

Piping more than ~10 steps is a bad idea, create intermediate objects with meaningful names to help with debugging. 

If there are multiple inputs or outputs, pipes are a bad idea. 

Directed graph with a complex dependency structure, bad. 

<br> 

### Other Margrittr Tools 

When you want to call a function for its side-effect and the function doesnt return anything useful it'll end a traditional pipe. the tee pipe `%T>%` works just like the normal pipe, but it returns the side effects. 

```{r}
library('magrittr')

rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()
```
^ ## NULL because the `%>%` forwarded the actual plot to str(), not the side effect
```{r}
rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```
^ `str()` worked!


