---
title: "R4DS Chapter 19 - Functions"
author: "Fisher Ankney"
date: "2/26/2018"
output: html_document
---

<br> 

#### Why Functions 

1. You can name them, making your code easy to understand
2. As requirements change, you can just update the function
3. Don't repeat yourself, (DRY), copying and pasting is bad

Consider writing functions whenever you find yourself copying and pasting code more than once. 

#### How to Create a Function

1. Name the function
2. List the arguements
3. Create the body

It's easier to make body code that does a job for you, and then figure out how to move it into a function than to create a function and try to make it work. 

#### Practice Problems
```{r}
# create a list with Inf and -Inf then map them to 1 and 2
y <- c(1:10, T, Inf, -Inf) 
y[is.infinite(y) & y > 0] <- 1 
y[is.infinite(y) & y < 0] <- 2 
y 

# create a rescale function 
rescale01 <- function(x) {   
  rng <- range(x, na.rm = T, finite = T)  
  (x - rng[1] / (rng[2] - rng[1]))         
}

# test function
rescale01(y)   
```
<br>

 
```{r}
# create a new list
x <- c(NA, 1:10)

# make a function giving the % of numbers missing in a vector
percent_na <- function(x) {
  mean(is.na(x))
}

# test function
percent_na(x)
```

<br> 

```{r}
# make a percent of total function
perc_total <- function(x) {
  x / sum(x, na.rm = T)
}

# test function
perc_total(x)
```

<br> 

```{r}
# turn this into a function:
# sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)

calc <- function(x) {
  y <- sd(x, na.rm = TRUE)
  z <- mean(x, na.rm = TRUE)
  y/z
}

# test function
calc(x)
```

<br>

```{r}
# make a funcion that counts number of common NA's 
v1 <- c(1,2,3,NA)
v2 <- c(NA,2,3,NA)

both_na <- function(x,y) {
  sum(is.na(x) & is.na(y))  
}

# test function
both_na(v1,v2)
```

<br>

#### Conditional Statements

Typical conditional if-else statement:
```
if (condition) {
  code executed when TRUE
} else {
  code executed when FALSE
}
```

```{r}
has_name <- function(x) {
  nms <- names(x) 
  if (is.null(nms)) {
    rep(FALSE, length(x))
  } else {
    !is.na(nms) & nms != ""
  }
}
```

Condition must evaulate to either True or FALSE, it its a vector you get a warning, if its NA you get an error. 

`|` and `&` are vectorized statements and should never be used in conditional statements. Use `||` and `&&` instead. 


`all()` - are all of these values true? <br>
`any()` - are any of these values true <br>
`indentical()` - are these exactly the same? <br>
`near()` - dplyr function, are these basically the same <br>

<br>

#### Multiple Conditions 

typical if-else if statement
```
if (this) {
  do that
} else if (that) {
  do something else
} else {
  do third thing
}
```

typical switch statement
```
function(x, y, op) {
  switch(op,
    plus = x + y,
    minus = x - y, 
    times = x * y,
    divide = x / y, 
    stop("unknown op!")
  )
}
```

<br> 

#### Practice Functions

Create a function that greets you based on the time of day 
```{r}
library('lubridate')

greetings <- function(time_now = lubridate::now()) {
  hour_now <- lubridate::hour(time_now)
  if (hour_now > 4 & hour_now < 12) {
    "good morning"
  } else if (hour_now >= 12 & hour_now <= 17) {
    "good afternoon"
  } else {
    "good evening"
  }
}

greetings()
```

Create a function that takes a single number and returns fizz if its divisable by 3, buzz if it's divisable by 5, and fizzbuzz if its divisiable by both. 
```{r}

fizzbuzz <- function(x) {
  if (x %% 3 == 0 & x %% 5 == 0) {
    "fizzbuzz"
  } else if (x %% 3 == 0) {
    "fizz" 
  } else if (x %% 5 == 0) {
    "buzz" 
  } else 
    "no fizzbuzz for you"
  }

fizzbuzz(1)
fizzbuzz(3)
fizzbuzz(5)
fizzbuzz(15)
```

Use `cut()` to simplify this set of nested if-else statements
```{r}
temp_original <- function(temp) {
  if (temp <= 0) {
    "freezing"
  } else if (temp <= 10) {
    "cold"
  } else if (temp <= 20) {
    "cool"
  } else if (temp <= 30) {
    "warm"
  } else {
    "hot"
  }
}
```

```{r}
temp = 2

temp_cut <- function(temp) {
  breaks <- c(-Inf, 0, 10, 20 , 30, +Inf)
  labels <- c("freezing", "cold", "cool", "warm", "hot")
  cut(temp, breaks, labels)
}

temp_cut(1)
```

<br> 

#### Function Arguements 

Two types of functions, the data to be computed on, and the details of the computation. 

`log()` the data is `x`, the detail is `base =` of the logarithm

Generally the data argument always comes first. 

Also space out the arguements with spaces after commas and between operators. That way it's easier to quickly skim what a function is doing. 

Common naming schemes: <br> 
`x`, `y`, `z`: vectors <br> 
`w`: a vector of weights <br> 
`df`: data frame <br> 
`i`, `j`: rows and columns <br>
`n`: length, number of rows <br>
`p`: number of columns <br>

<br> 

#### Checking Values

R uses vector recycling, meaning that you'll get screwed over if you assume that vectors will be the same length but they are not. 

```{r}

wt_mean <- function(x, w) {
if (length(x) != length(w)) {
  stop("`x` and `w` must be the same length", call. = FALSE)
}
  sum(w * x) / sum(w) 
}

wt_mean(c(1:4), c(4:8))
wt_mean(c(1:4), c(4:7))
```

<br> 

Be careful about being too careful, functions are supposed to improve efficiency, not suck it away. 

Many functions let you input an arbitrary number of values, the `...`. 

```{r}
commas <- function(...) {
  stringr::str_c(..., collapse = ", ")
}

commas(letters[1:10])
```

<br> 

#### Return 

`return()` breaks circuits early, returning whatever you wish. Useful for early returns before complex functions or missing arguements. 

```
complex_function ,- function(x, y, z) {
  if (length(x) == 0 || length(y) == 0 {
    return(0)
  } else {
  
  # complex stuff
  
  }
}
```

<br> 

#### Writing pipeable functions

It's important to think about return values if you want to write a pipeable function. Transform the object into a dataframe inside the function to pass it on to other dplyr and tidyr functions. 







