---
title: "R4DS Chapter 20 - Vectors"
author: "Fisher Ankney"
date: "2/27/2018"
output: html_document
---

<br> 

### Introduction

Vectors are the objects that underly dataframes and tibbles. Most resources start with vectors and work their way up to dataframes, but this book does it in the opposite direction because dataframes are immediately useful. This chapter will use a handful of functions from the package `purrr` in the tidyverse. 

```{r, message=FALSE}
library('tidyverse')
```

There are two types of vectors: 

**Atomic Vectors** - Homogeneous<br> 
- Logical <br> 
- Integer <br> 
- Double <br> 
- Character <br> 
- Complex <br> 
- Raw <br> 

**Lists** - Heterogeneous, can contain multiple vector types<br> 

`NUll` = absence of a vector 
`NA` = absence of a value in a vector 

<br> 

`typeof()` - determines type of vector
```{r}
typeof(letters)
typeof(TRUE)
```

<br> 

`length()` determines the length of a vector
```{r}
length(letters)
length(TRUE)
```

<br> 

Augmented Vectors are additional metadata in the form of attributes that compliment a vector. Factors are built on top of integer values, dates and date-times are built on top of numeric vectors, and data frames and tibbles are built on top of lists.

<br> 

### Important Atomic Vectors 

<br> 

**Logical Vectors** are the most simple type of atomic vector with three possible values: TRUE, FALSE, and NA. 

```{r}
1:10 %% 3 == 0
```

<br> 

**Numeric Vecors** are integers and doubles. Numbers are doubles by default, to make a number an integer, plage an `L` after the number. Doubles are approximations representing floating point numbers that can not always be precisely represented with a limited amout of emmory. 

```{r}
typeof(1L)
``` 

Example of the floating point problem: 
```{r}
x <- sqrt(2) ^ 2
x

y <- x - 2
y
```

<br> 

Use `dplyr::near()` to overcome some of the challegnes that the floating point problem presents. Checking for special values is as easy as these four functions: `is.finite()`, `is.infinite`, `is.na()`, `is.nan()`. 

<br> 

**Character Vectors** are the most complex type of atomic vectors. Each element of a character vector is a string, and strings contain arbitrary amounts of data. 

<br> 

### Coercion 

<br> 

**Explicit Coercion** <br> 
`as.logical()` <br> 
`as.integer()` <br> 
`as.double()` <br> 
`as.character()` <br> 

Whenever you use these you should see if you could have fixed the
problem somewhere upstream. `col_types` from the readr package may come in handy. 

<br> 

**Implicit Coercion**
A vector is used in a specific context that expects a certain type of vector. For example, when you use a logical vector with a numeric summary function, or when you use a double vector where an integer is expected. 

Logical to integer -  
```{r}
x <- sample(20, 100, replace = TRUE)
y <- x > 10
sum(y) # how many are greater than 10 
mean(y) # what proportion are greater than 10 
```

<br> 

Integer to logical - 
```{r, eval=FALSE}
if (length(x)) {
 # do something
}
```

<br> 

It's important to remember that atomic vectors cannot mix their types, the most complex type wins. 
```{r}
typeof(c(TRUE, 1L, "a"))
```

<br> 

### Purrr Test Functions

<br> 

`is_logical()` - Logicals TRUE <br>
`is_integer()` - Integers TRUE <br> 
`is_double()`  - Doubles TRUE <br>
`is_numeric()` - Int and Doubles TRUE <br> 
`is_character()` - Characters TRUE <br> 
`is_atomic()`  - Logical, Numeric, Character TRUE <br>
`is_list()`    - Lists TRUE <br>
`is_vector()`  - All of the above TRUE <br> 

<br> 

### Recycling

R implicitly coerces the length of vectors, this is called vector recycling. The shorter vector is repeated until it matches the length of the longer vector. 

Recycling Example:
```{r}
1:10 + 1:2
```

<br> 

Tidyerse throws errors at you because it's easy for this silent coercion to really screw up you analysis. If you want to recycle anything within the tidyverse use the `rep()` function. 

<br> 

Base R: 
```{r}
data.frame(x = 1:4, y = 1:2)
```

<br> 

Tidyverse:
```{r, eval=FALSE}
tibble(x = 1:4, y = 1:2)
# Error 
```

<br> 

Tidyverse Reps:
```{r}
tibble(x = 1:4, y = rep(1:2, each = 2))
```

<br> 


### Naming and Subsetting

<br>

All vectors can be named with `c(x = 1)` or `purr::set_names(1:2, c("a", "b"))`. Naming is useful for subsetting. 

Subsetting `x[]` is called with square brackets. There are four types of things that you can subset a vetor with.

`x[c(1,5,8)]` - positive integers that keep the elements at those positions <br>
`x[c(-1,-4)]` - negative integers that drop the elements at those positions <br> 
`x[!is.na(x)]` - logical vector keeps all values that return true <br> 
`x[c("xyz")]` - subsetting a previously named vector <br> 
`x[,,]` - returns the complete vector, useful with matrices <br> 

There is an important variattion of `[` called `[[`. `[[` only ever extracts a signel element and always drops the name. Just remember, one bracket calls mutliple elements, multiple brackets call one element. 


### Lists 

Lists are a step up in complexity from atomic vectors because they can contain other lists. They can also contain a mix of objects. 

```{r}
x <- list(1,"a", TRUE, list(4,5))
str(x)
```

<br> 
```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1,-5))

str(a[1:2])
str(a["d"])

str(a[[1]])
str(a[[4]])

str(a$d)
```

<br> 

### Attributes

Any vector can contain arbirary additional metadata through its atributes. Attributes can be thought of as a named list of vectors that can be attached to any object. 

`attr()` <br> 
`attributes()` <br>

Vectors with attributes are augmented vectors, augmented vectors include 
- Factors <br> 
- Dates <br> 
- Date-Times <br> 
- Tibbles <br> 


That's all for chapter 20 notes. 