---
title: "R4DS Chapter 21 - Iteration"
author: "Fisher Ankney"
date: "2/28/2018"
output: html_document
---

<br> 

### Introduction
 
Reducing the duplication in your code by creating functions is useful in software development and data science. It's easier to see the intent of your code, it's easier to change, and you're likely to have fewer bugs. Another tool for reducing duplication is iteration. Iteration helps you when you need to do the same thing to multiple inputs. Imperative programming and functional programming are two important iteration paradigms. On the imperative side you have for loops and while loops, they're very explicit. Functional programming offers tools for extracting out the duplication found in imperative programming, so a common for loop pattern will get it's own function. 

```{r, message=FALSE}
library('tidyverse')
```

<br> 

### For Loops 

Imagine we have this simple tibble: 
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

<br> 

We want to compute the median of each column, you could do this with copy-and-paste:
```{r, eval=FALSE}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

<br>

Nah that's stupid, use a for-loop:
```{r}
output <- vector("double", ncol(df)) # output
for (i in seq_along(df)) {            # sequence
  output[[i]] <- median(df[[i]])      # body 
}

output
```

<br> 

Every for loop has three components 

**Output** - `output <- vector("double", ncol(df))` - Before you start the loop you must always allocate sufficient space for the output. This is very important for efficiency: if you grow the for loop at each iteration using `c()` your for loop will be very slow. A general way of creating an empty vector of given length is the `vector()` function. It has two arguements: the type of vector, and the length of the vector. 

**Sequence** - `i in seq_along(df) {` - Determines what to loop over, each run of the loop will assign i to a different value from seq_along(df). `seq_along()` is a safe version of `1:length(l)` because if you have a zer length vector, `seq_along()` resonds correctly. 

**Body** - `output[[i]] <- median(df[[i]])` - This is the code that does the work. It's run repeatedly, each time with a different value for `i`. The first iteration runs `output[[1]] <- median(df[[1]])`, the second runs `output[[1]] <- median(df[[1]])`, and so on. 

<br> 

### Practice 

Write a loop to compute the mean of every column in mtcar 
```{r}
head(mtcars)

mtcars_out <- vector("double", ncol(mtcars))

for (i in seq_along(mtcars)) {
  mtcars_out[[i]] <- mean(mtcars[[i]])
}

mtcars_out
```

<br> 

Determine the type of column in nycflights13::flights
```{r}
library('nycflights13')

flights_out <- vector("list", ncol(flights))

for (i in seq_along(flights)) {
  flights_out[[i]] <- class(flights[[i]])
}

head(flights_out)
```

<br> 

Compute the number of unique values in each column if `iris`
```{r}
head(iris)

iris_unique <- vector("double", ncol(iris))
 for (i in seq_along(iris)) {
   iris_unique[[i]] <- length(unique(iris[[i]]))
 }

iris_unique
```

<br> 

Write a for loop that prints the lyrics to "alice the camel" 
```{r}
num_humps <- c(4,3,2,1,"no")

for (i in num_humps) {
  cat(str_c("Alice the camel has ", rep(i, 3), " humps",
            collapse = "\n"), "\n")
  if (i == "no") {
    cat(str_c("Now alice is a horse", "\n"))
    } else {
      cat(str_c("So go Alice go", "\n", "\n"))
    }
}
```

<br> 

Convert the nursery rhyme "ten in the bed" to a function generalizing any number of people in any structure. 
```{r}
in_the_bed <- function(x, struct) {

    y <- vector("double", length = x + 1)
 
     for (i in y) {
       if (x != 0) {
      cat("There were", x, "in the", struct, "\nAnd the little one said", "\n'Roll Over! Roll Over'", "\nSo they all rolled over and one fell out", "\n\n")
      x <- x - 1
       } else {
      cat("There was", x, "in the", struct, "\nAnd the little one said", "\nAlone at last!")
     }
  }
}

in_the_bed(4, "Walmart")
```

<br> 

Convert the song "99 botles of beer on the wall" to a function. Generalise to any number of vessels containing any liquid on any surface. 
```{r}
on_the_wall <- function(w, x, y, z) {

  l <- vector("double", length = w + 1)
  
  for (i in l) {
    cat("There are", w, x, "of", y, "on the", z, "\n", w, x, "of", y, "\n Take one down, Pass it around\n", w-1, x, "of", y, "\n\n")
    w <- w - 1
    }
}

on_the_wall(4, "jugs", "vitamin water", "floor")
```
Close enough. 

<br> 

### For Loop Variation 

1. Modify and existing object, instead of creating a new one <br> 
2. Looping over names or values, instead of indices <br> 
3. Handling outputs of unknown length <br> 
4. Handling sequences of unknown length <br> 

<br> 

**Modifying an existing object** - Remember this from functions? 
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

<br> 

To solve it with a for loop, we need to think about the three components:

1. Output: we already have the output. it's the same as the input <br> 
2. Sequence: we can think about a data frame as a list of columns, so we can iterate over each column with seq_along(df) <br>
3. Body: apply rescale01() <br> 

This gives us: 
```{r, eval=FALSE}
for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}
head(df)
```

<br>

**Looping Patterns** - There are three basic ways to loop over a fector. so far we've seen the most general `for (i in seq_along(xs))`, and extracting the value with `x[[i]]`. There are two other forms. Loop over the elements `for (x in xs)` This is the most useful if you only care about side-effects. like plotting or saving a file, because it's difficult to save the output efficiently. Looping over the name `for (nm in names(xs))`. This gives you name, which you can use to access the values with `x[[nm]]`. This is useful if you want to use the name in a plot title or file name. If your're creating a named output, make sure to name it like so: 
```{r, eval=FALSE}
x <- 1

results <- vector("list", length(x))
names(results) <- names(x)
```

<br>

Iterating over the numeric indices is the most general form, becuse given the position you can extract both the name and the value: 


```{r, eval=FALSE}

for (i in seq_along(x)) {
  name <- names(x)[[i]]
  value <- x[[i]]
}

```

<br> 

**Unknown Output Length** - Never program a growing vector! A better solution is to save the results in a list, then combine into a single vector after the loop is done. 

```{r}
means <- c(0,1,2)

out <- vector("list", length("means"))

out

for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}

str(out)

str(unlist(out))
```

<br> 

`unlist()` flattens a list of vectors into a single vector. Another useful function is `paste(output, collapse = "")` which will combine character vectors into a signle string. `dplyr::bind_rows(output)` combines output lists into a big dataframe effectively. 

<br>

**Unknown Sequence Length** - Sometimes you don't know how long the input sequence should run for. That's when you use a while loop. A while loop is more general than a for loop, because you can write any for loop as a while loop, but you can't write any while loop as a for loop. 

This for loop:
```{r, eval=FALSE}
for (i in seq_along(x)) {
  # body 
}
```

can become this while loop:
```{r, eval=FALSE}
i <- 1 

while( i <= length(x)) {
  # body
  i <- i + 1
}
```

<br> 

Here's how to create a while loop to find how many tried it takes to get three heads coin flips ina row:
```{r}
flip <- function() {
  sample(c("T", "H"), 1)
}

flips <- 0 
nheads <- 0 

while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0 
  }
  flips <- flips + 1
}

flips
```

<br> 

While loops are most often used for simulations, which are out of the scope of this book but seem really awesome. 

<br> 

### Practice!

write a loop that will load a directory full of csv files into a signle dataframe 
```{r}

files <- dir("data/", pattern = "\\.csv$, full.names = TRUE")

df <- vector("list", length(files))

for (fname in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}

df <- bind_rows(df)
```

<br> 

Write a function that prints the mean of each numeric column in a dataframe along with its name. 
```{r}
show_mean <- function(df, digits = 2) {
  # Get max length of any variable in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
                format(mean(df[[nm]]), digits = digits, nsmall = digits),
                sep = " "),
          "\n")
    }
  }
}
show_mean(iris, 2)
```

<br> 

### For Loops vs Functionals 

For loops are not as important in R as they are in other languages because R is functional. That means that usually for loops are wrapped up in functions instead of called directly. Imagine you want to compute the mean of every column in this dataframe:

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

<br> 

You could do it with a for loop: 
```{r}
output <- vector("double", length(df)) 

for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}

output
```

<br> 

You realise that you're goingot want to comput the means of every column pretty frequently, so you extract it out into a function: 

```{r}
col_mean <- function(df) {
  output <- vector("double", length(df))
  for ( i in seq_along(df)) {
    output[[i]] <- mean(df[[i]])
  }
  output
}
```

<br> 

But then you think it'd also be helpful to be able to compute the median, and the standard deviation, so instead of copying and pasting the `col_mean()` function and replace the `meean()` with `median()` and `sd()`, let's just create a composie function. 

```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

col_summary(df, median)
col_summary(df, mean)
col_summary(df, sum)
```

<br> 

The idea of passing a function to another function is extremely powerful. The rest of the chapter is focused on learning `purrr` package functions that are meant to replace `apply()`, `lapply()`, and `tapply()` in base R. The goal of using purrr instead of for loops is to allow you to break common list manipulation challenges into independent peices. 

- How can you solve the problem for a signle element of the list? Once you've solved that problem, purrr takes care of generalizing the solution. 

- If you're solving a complex problem, how can you break it down into bite sied peices? With purrr,you get alots of small pieces that you can compose together with the pipe. 

<br> 

### The Map Functions

The pattern of looping over a vector, doing something to each element and saving the result is so common that purr provides a family of functions to do it for you, there is one function for each type of output. 

`map()` - makes a list <br> 
`map_lgl()` - makes a logical vector <br> 
`map_int()` - make an integer vector <br> 
`map_dbl()` - makes a double vector <br> 
`map_chr()` - makps a character vector <br> 

Each function takes a vector as input, applies a function to each piece, then returns a new vector thats the same length and has the same names as the input. the type of vector is determined by the suffix to the map function. 

Using the map function to perform the same computation as the last for loop: 
```{r}
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sum)
```

<br> 

The focus is on the verb, not the noun. This is even more apparent when you use a pipe: 
```{r}
df %>% map_dbl(mean)
```

<br> 

All purrr functions are implemented in C. This makes them faster at the expense of readability. The second arguement, .f, the function to apply, can be a formula, a character vector, or an integer vector. `map()` uses ... to pass along additional arguements to .f each time it is called. Map also preserves names. 

```{r}
map_dbl(df, mean, trim = 0.5)
```

```{r}
z <- list( x = 1:3, y = 4:5)
map_int(z, length)
```

<br> 

There are af few shortcuts you can use within the map functions. Imagine wanting to fit a linear model to each group in a dataset. This example splits up `mtcars` into three peices (one for each value of cylinder) and fits the same linear model to each piece: 

```{r}
models <- mtcars %>% 
  split(.$cyl) %>%
  map(function(df) lm(mpg ~ wt, data = df))

models
```

The syntax for creating an anonymous function is verbose in R, so a one-sided formula is a convenient shortcut: 

```{r}
models <- mtcars %>% 
split(.$cyl) %>% 
map(~lm(mpg ~ wt, data = . ))
```

`.` is used as a pronoun in this case, it refers to the current listed element, in the same way that `i` refers to the current index in the for loop. 

Extracting the Rsquared of a model is easy: 
```{r}
models %>% 
  map(summary) %>%
  map_dbl(~.$r.squared)
```

but extracting named components is a common operation so purr provides an even shorter shortcut: you can use a string: 

```{r}
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```

You can also yuse an integer to select elements by position: 

```{r}
x <- list(list(1,2,3), list(4,5,6), list(7,8,9))
x %>% map_dbl(2)
```

<br> 

### Practice! 

Write code that uses the map function to compute the mean of every column in mtcars: 
```{r}
mtcars %>% 
  map_dbl(mean)
```

determines the type of column in each nycflights13::flights
```{r}
library('nycflights13')

flights %>% 
  map(class)
```

computes the number of unique values in each column of iris
```{r}
iris %>%
  map(unique) %>%
  map_int(length)
```

rewrite map(x, function(df) lm(mpg ~ wt, data = df)) to eliminate the anonymous function
```{r, eval=FALSE}

x %>% 
  map(~lm(mpg ~ wt, data =  df))


```

<br> 

### Dealing With Failure

Chances are that an operation in a map function will fail at some point. How do you ensure that one bad apple (failure) doesnt ruin the whole map function? `safely()` that's how. `safely()` will cause a list to return with two elements, result and error. 

```{r}
x <- list(1,10,"a")
y <- x %>% map(safely(log))
str(y)
```

<br>

### Mapping Over Multiple Arguements

So far we've mapped along a signle input, but often you'll have multiplerelated inputs. that's the job of `map2()` and `pmap()`. So imagine you want to simuate some random normals with diferent means. You know how to do that with `map()`
```{r}
mu <- list(5,10,-3)
mu %>% 
  map(rnorm, n=5) %>%
  str()

```

<br> 

but what if you want to vary the standard deviation as well? That's when you use `map2()` to iterate over two vectors in parallel. 

```{r}
mu <- list(5,10,-3)
sigma <- list(1,5,10)

map2(mu, sigma, rnorm, n = 5) %>% str()
```

<br> 

Note that arguements that vary for each call come before the function, and arguements that are the same for each call come after the function. Effectively, map2 is making the three functions `rnorm(5, 1, n = 5)`, `rnorm(10, 5, n = 5)` and `rnorm(-3, 10, n = 5)`. Here's what this looks like with a base for loop: 

```{r}
maps <- function(x, y, f, ...) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]], y[[i]], ...)
  }
  out
}
```

<br> 

In this sense, you can imagine a multitude of map functions for each number of variables. This is wher `pmap()` comes in handy. 
```{r}
n <- list(1,3,5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>%
  str()
```

<br> 

