---
title: "R4DS - Reading Data with readr"
author: "Fisher Ankney"
date: "February 13, 2018"
output: html_document
---
This is the second of a six part series summarsing the core concepts of Hadley Wickham's R for Data Science: <http://r4ds.had.co.nz/>. In the previous post, I've written about the R4DS chapters covering data visualization using ggplot2, you can find that post [here](site.baseurl/).

in this post, I'll summarize importing and exporting data with the readr library. While importing data may sound trivial, it can become a hassle when you're faced with parsing date-times, unusual file-types, non-typical encodings, or foreign language syntax. 

For more useful resources on importing data, check out these links:

-
-
-
-


<br  />

#### Load Libraries
```{r, message=FALSE}
library('tidyverse') # used for many amazing things, includes readr library
library('readr')     # used to read in a variety of file types
library('readxl')    # used for reading in microsoft excel files
```

You may also be interested in the following libraries, however I won't cover them in this blog post - 

```{r, eval=FALSE}
library('haven')       # used for SPSS, SAS, and Strata input files
library('DBl')         # used for database files, must combine with specific backend
library('RMySQL')      # used for MySQL backend (with DBl)
library('RSQLite')     # used for SQLLite backend (with DBl)
library('RPostgreSQL') # used for PostgreSQL (with DBl)
```



### Reading Data
local paths
setwd()
getwd()
local data types
online paths

### Parsing Data
columns
date-times
all that nonsense

### Writing Data 
goes to outputs
scripts save in scripts
