---
title: "R4DS - Tidying Data with tidyr"
author: "Fisher Ankney"
date: "February 21, 2018"
output: html_document
---

This is the fourth installment of a six-part series summarizing concepts from Hadley Wickham's textbook, [R for Data Science](http://r4ds.had.co.nz/). In the previous [blog post](link) I abridged the book’s chapter that covers data wrangling with dplyr. 

In this installment, I'll be summarizing the tidyr package and tidy data practices, as taught in R for Data Science [Chapter 12](http://r4ds.had.co.nz/tidy-data.html). There are four components to the chapter and this summary blog post, the first being "what constitutes tidy data"?. The second and third component deal with correcting the two typical formatting issues of untidy data, and the fourth component dives into how to properly represent missing values in a dataset. 

For more resources on tidyr, reference these links:

- [CRAN tidyr Documentation](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
- [STHDA tidyr Tutorial](http://www.sthda.com/english/wiki/tidyr-crucial-step-reshaping-data-with-r-for-easier-analyses)
- [Journal Of Statistical Software: Tidy Data by Hadley Wickham](http://vita.had.co.nz/papers/tidy-data.pdf)

<br  />

### Libraries
```{r, message=FALSE}
library('tidyverse')      # includes tidyr
library('tidyr')          # tidyr specifically
```

<br  />

### Primary Functions

```{r eval=FALSE}
spread()        # spreads a key-value pair across multiple columns
gather()        # takes multiple columns and collapses into key-value pairs
separate()      # single column into multiple columns, defined seperator
unite()         # multiple columns into single column, defined seperator
fill()          # fills missing values using the previous entry
complete()      # takes a set of columns and finds all unique combos
```

<br  />

### Tidy Data

There are three rules to tidy data -

1. Each column is a variable <br  />
2. Each row is an observation <br  />
3. Each value has it's own cell <br  />

<insert tidy data image here> 

Tidy data is an easy and consistant way of storing data that makes further analytical steps more simple. Datasets that follow the three tidy data rules allow for R's vectorized nature to work it's magic. Other packages in the tidyverse, such as dplyr and ggplot2, expect datasets to be tidy when applying functions such as *mutate( )* and *ggplot( )*. Datasets may grow in size during the tidying process, but breaking down rows and columns into single observations and variables drastically simplifies the dataset for future analytics.

Here's an example of a tidy dataset - 

```{r}
table1
```

<br  />

Here's an example of an untidy dataset - 

```{r}
table2
```

<br  />

Notice the differences between the two tables above; in table 1 the variables (columns) are country, year, cases, and population. In table 2 the variables are country, year, type and count. In table 2 the data is untidy; the observation, a specified country in a specified year, is spread among multiple rows. This is because population and cases are actually variables that should be column names, not values! 

<br  /> 

### Spread and Gather

Spread and gather are two functions in the tidyr package that address these issues. While these functions are easy to use and understand, it's often quiet hard to determine what consitutes a discrete observation, and what constitutes a discrete variable. It takes time and practice to properly make these choices, but in general, I aim to break down observations and variables into the smallest and most easily retrievable chunks possible. 

<br  />

#### One Observation Across Multiple Rows

The oppsoite of gather, used when an observation is scattered across mutliple rows. table2 is an example, an observation is a country in a year but each observation is spread across multiple rows. 

```{r}
table2 %>%
spread(key = type, value = count)
```

<br  /> 

table 2's type column contains two different variables, cases and pop
we need to spread these into two new column heads, so spread(key = type, 
and the associated data is column count, or value = count). thats it!


<br  />

#### One Variable Across Multiple Columns

```{r}
table4a
```



```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

<br  />

'gather' columns titled `1999`, and `2000`, as these are values, 
not variables. 

'key' a new column called 'year' to put 1999 and 2000 into. 

values associated with the new column 'year' need a new home
we now call them 'cases'. they still match with the intended years 


<br  />


### Seperate and Unite 

#### One Column With Multiple Variables

```{r}
table3
```

<br  />

```{r}
table3 %>%
  separate(rate, into = c("cases", "population", sep = "/"))
```

<br  />

pulls one column apart into multiple columns, default is by a 
non alphanumeric chr, but put in the sep variable if you want 

<br  />

#### Single Variable Among Multiple Columns 

the inverse of seperate, combines multiple columns into a single column

```{r}
table5
```


```{r}
table5 %>%
  unite(full_year, century, year, sep = "")

```

declare a new column name, and unite two defined old variables by 
the indicated seperator. 



<br  />
<br  />
<br  />



### Fill and Complete 

```{r eval=FALSE}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra="drop")
# drops that extra value completely 

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill ="right")
# fills the row by putting an NA on the right 
```


#### Fill 

```{r eval=FALSE}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment %>% 
  fill(person)

# fill() will carry forward the last value for that column
```

<br  />

#### Complete 
```{r, eval=FALSE}
complete()
```




that is tidyr, super useful package! summary summary summary

Until next time, <br  />
\- Fisher