---
title: "Unpacking the Tidyverse - ggplot2"
author: "Fisher Ankney"
date: "January 19, 2018"
output: html_document
---

<br> 

### Introduction 

<br> 

This is the first of eight installments of my *Unpacking the Tidyverse* series. Each installment focuses on one of the eight core packages in Hadley Wickham's tidyverse. Instructions given in each post are mainly derived from Hadley's textbook, [R for Data Science](http://r4ds.had.co.nz/), and CRAN package documentation. This installment of *Unpacking the Tidyverse* focuses on the graphing package, ggplot2. The next installment focuses on the readr package and can be found [here](link).

The ggplot2 package is a powerful data visualization tool based on the grammer of graphics: an idea that any plot can be described using seven distinct parameters. In this blog post, I run through these parameters attempting summarize them in an easy-to-access resource for some of the most common tasks in data visualization. 

<br> 

```{r message=FALSE}
library('ggplot2')       
library('RColorBrewer')  
library('viridisLite')  
```

<br> 

### Template

<br> 

```
ggplot( ***dataset*** ) +            
geom_***geometry***(              
  mapping = aes( ***variables*** ),    
  stat = ***statistical transformation***,          
  position = ***position adjustment***        
) + 
coord_***coordinate system****() +          
facet_***facet declaration***()              
```

<br> 

Not all of these parameters must be defined for every ggplot call, but by defining the dataset, plot geometry, variables, statistical transformations, position adjustments, coordinate system and facets, you can create almost any plot with ggplot2. Of course, there are additional functions that affect the aesthetics of the plot, such as annotations, titles, axes, and legends. Let's briefly cover all seven of these parameters, discussing the commands that I most commonly find useful for each of them. 

<br  /> 

### Dataset

<br> 

```{r}
ggplot(mpg)
```

<br> 

Just calling `ggplot()` with an assoicated dataset will create a canvas for your graphic, not an actual plot. This is because R doesn't know what variables you want plotted from that dataset, or how you want them plotted. In order to create a basic plot, you'll have to provide information on the plot's geometry. 

<br  />

### Geometries

<br> 

```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = cyl))
```
     
<br> 

Plot geometries are the most important parameter, and the only parameter other than `ggplot()` that must be defined in every single visualization. There are hundreds of geometries avaliable for your use, courtisy of the R community; check out http://www.ggplot2-exts.org/ for more information. Here's a list of the handful of geometries, mappings, and aestetics that I use most often. 

##### Geometries    
* **geom_density( )** - 1 variable gaussian distribution
* **geom_point( )** - 2, 3, or 4 variable scatterplot 
* **geom_smooth( )** -  2 or 3 variable line plot
* **geom_bar(stat="identity")** -  2 or 3 variable bar plot
* **geom_hex( )** - 2 variable distribution
* **geom_tile( )** - 3 variable tile plot 
* **geom_map( )** - 2, 3, or 4 variable geospatial

##### Mappings 
* **aes( x = )** - x-axis location
* **aes( y = )** - y-axis location
* **aes( ..., color = )** - color dependent on variable's value
* **aes( ..., linetype = )** - linetype dependent on variable's value
* **aes( ..., size = )** - size of object dependent on variable's value

##### Aestetics 
* **size =** size of a point, declared by an integer
* **linetype =** 0 (blank), 1 (solid), 1 (dashed), 3 (dotted), 4 (dot-dash), 5 (long-dash), 6 (two-dash)
* **weight =** (line thickness) integer
* **color =** (outline) string
* **fill =** (inside) string
* **alpha =** (transparency) double from 0 to 1    

<br  />

### Statistical Transformation

<br> 

Statistical transformations are most useful with `geom_bar`. Each geometry has an associated transformation, and most of the time this transformation is useful and should not be changed. The default transformation associated with `geom_bar` is `"count"`, or the total number of occurances of each variable. Sometimes you want a bar chart to visualize something else, like the unique occurances of each variable, or a different variable all together. These three transformations allow you control over bar charts to visualize your data in entirely new ways.      
    
* **geom_bar(aes(...), stat = "count")** - visualise the number of entires in a variable, n()    
* **geom_bar(aes(...), stat = "identity")** - visuaise the variable, not the count    
* **geom_bar(aes(...), stat = "unique")** - visualise only unique components of the variable    

<br  />

### Position Adjustment

<br> 

When geometric data occupies the same space, it is useful to use a position adjustment to better understand the visualisation. I really only find adjusting the positioning of data useful when I'm working with bar charts and scatterplots.   
    
* **geom_point(aes(...), position = "jitter")** - adds a small amount of noise to better view overlapping points     
* **geom_bar(aes(...), position = "dodge")** - arranges bar elements side-by-side    
* **geom_bar(aes(...), position = "fill")** - arranges bar elements on top of each other, normalizing     height    

<br> 
   
### Coordinate Systems

<br> 

```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  coord_flip()  
```

<br  />     

Cartesian coordinates are the go-to for 95% of my visualisations, but the coord_ function does more than call different coordinate systems. Often it's useful to fix the x-y ratio, or flip the axis of bar charts. When your working with exponential data, you may want to transform an axis to a logarithmic system. These useful functions and many more can be done by the coord_ function.    
    
* **coord_fixed(ratio, xlim, ylim)** - fixing the aspect ratio between x and y
* **coord_flip(xlim, wlim)** - flipping the axes 
* **coord_polar(theta, start, direction)** - coverting cartesian to polar coordinates
* **coord_trans(xtrans, ytrans, limx, limy)** - transform cartesian coordiantes, use with log functions    
* **coord_map(projection, orientation, xlim, ylim)** - mapproj package projections 
   
<br>   

### Facets 

<br> 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
  
<br> 

I don't use facets that much; to be honest I find them pretty distracting. If you're a fan of contrasting minerature plots, you might find these faceting options useful.    

* **facet_grid(x ~ y, labeller, scales)** - facets the display into rows and columns based on the two given variables ("."" placeholder)    
* **facet_wrap(~ x, labeller, scales, nrow, ncol)** - wraps facets into a rectangular layout    

<br  /> 
     
### Labels

<br> 

```{r}
ggplot(mpg) + 
  labs(
    title = "title", 
    subtitle = "subtitle",
    caption = "caption",
    xlab = "x-label", 
    ylab = "y-label"
  )
```
      
Labels are helpful for properly communicating your visualisation. Including these five labels with your plots can help clarify your topic, variables, and sources.    

<br  />

### Annotations 

<br> 

Annotations are another feature I don't use too often, but I'll include them anyways. I think annotating individual observations is a mistake, it clutters the graphic and takes away from the effectiveness of the visualisation. Packages like plotly and shiny do a better job of creating interactive graphics if you want individual labels. If you, however, insist on creating a ggplot with annotations, these might be useful.     
      
* **geom_text(aes(label = "text here"), vjust = , hjust = )** - places a summary annotation within the graphic, according to vjust and hjust     
* **geom_hline(yintercept = , size = , color = , linetype)** - creates a horizontal reference line through the plot according to yintercept     
* **geom_vline(xintercept = , size =, color =, linetype =)** - creates a verticle reference line through the plot according to xintercept     
* **geom_rect(aes(xmin = , xmax = , ymin = , ymax = ))** - creates a rectangle to box in and highlight interesting data    
* **geom_segment(aes(x = , y = , xend = , yend = ,) arrow = )** - creates a line segment within the plot, it can be an arrow     
     
<br  />
 
### Scales    
 
<br> 

Scales are automatically set when you create a ggplot, but sometimes it's useful to alter the color scheme, legend, or axes. The naming sceme for these functions is scale_ followed by the name of the aesthetic then another _, and then the name of the scale.

##### Aestetic Names
     
  * **\_x\_**    
  * **\_y\_**      
  * **\_alpha\_**      
  * **\_color\_**           
  * **\_fill\_**         
  * **\_linetype\_**         
  * **\_shape\_**          
  * **\_size\_**        

##### Scale Names     
      
  * **\_discrete** - maps discrete variables to visual values      
  * **\_continous** - maps continuous variables to visual values
  * **\_identity** - uses data values as visual values 
  * **\_manual** - values = c( ), maps discrete values to manually chose visual values 
  * **\_gradient** - creates a two color gradient, low - high
  * **\_gradient2** - creates a diverging two color gradient, positive - negative
  * **\_gray** -creates a gradient in gray
  * **\_brewer** - brewer( palette = " "), calls RColorBrewer palettes    
     
<br  />

### Axes  

<br> 

```{r eval=FALSE}
scale_x_continuous(breaks = seq(0, 10, by = 1), labels = c(1:9, "ten")) 
# define breaks by sequence, define labels of breaks

scale_x_discrete(breaks=c("ctrl", "trt1", "trt2"),            # defined tick marks
                 labels=c("Control", "Treat 1", "Treat 2"))   # new names
                 
expand_limits(x = c(0,8), y = c(0,8))                         # expand the limits of the graph to visualise specific values

scale_y_reverse()     # Reverse y-axis direction (zero on top)

element_blank()       # Hiding axis labels 

theme(axis.title.x = element_text(face="bold", color="red", size=20),  # change axis labels, font and style
      axis.text.x = element_text(angle=90, vjust=0.5, size=16))        # change tick mark label, font, style

theme(panel.grid.major=element_blank(),  # hiding major gridlines
      panel.grid.minor=element_blank())  # hiding minor gridlines

library(scales)   # changing the y-axis to log 10
scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))
```

<br> 

Scales can be used to alter axes in ggplot2, it's a fairly straight forward process. These are some commented examples for fixing common behavioral issues you may have with your axes.   

<br  />     

### Legends     

<br> 

```{r eval=FALSE}

theme(legend.position = "left")    # legend to the left of plot
theme(legend.position = "top")     # legend above plot
theme(legend.position = "bottom")  # legend below plot
theme(legend.position = "right")   # the default
theme(legend.position = "none")    # no legend will be generated

geom_point(aes(...), show.legend=FALSE) # don't include this geom in the legend    

theme(legend.title = element_blank()) # hides the legends title

# changing the order of items in your legend, scale_ function is dependent upon the type of plot
scale_fill_discrete(breaks=(c("item1", "item2", "item3")))

# simply reverse the current item display in thelegend
scale_fill_discrete(guide = guide_legend(reverse=TRUE))

# manually alter color, break, label, and name of legend items.
scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"), 
                  name="Experimental\nCondition",
                  breaks=c("ctrl", "trt1", "trt2"),
                  labels=c("Control", "Treatment 1", "Treatment 2"))
```
    
<br> 

Legends can be a bit harder to wrangle than axes. Heere are examples to several common problems I have with legends.

<br  />

### Colors 

<br> 


Some of my favorite preset colors include 

* thistle
* slateblue2
* orchid
* midnightblue
* deepskyblue
* lightslateblue    

The RColorBrewer package also has some great color palettes. 
```{r eval=FALSE}
RColorBrewer::display.brewer.all()
```

<br> 

INSERT VIRIDISLITE EXAMPLES TOO

### Themes      

<br> 

Sometimes you don't want to go through the effort of creating your own theme, so ggplot2 has some built in. Here are four basic themes that are excellent for exploratory analysis or creating a solid foundation to build off.   

* **theme_bw()** - A while background with gridlines
* **theme_grey()** - the default theme 
* **theme_classic()** - A white background with no gridlines
* **theme_minimal()** - mystery theme, check it out! 

<br  />    

### Conclusions  

<br> 

This is just a fraction of what ggplot2 can do, but with this infromation I think you should be able to solve 80% of your graphing needs. Again, the references listed at the beginning of this write up are excellent and worth checking out. If you want to learn more about R for Data Science written by Hadley Wickham, it's free to read at <http://r4ds.had.co.nz/>. If you're too lazy to read the entire book, check out the other posts in my R for Data Science Summary Series.    

For more useful resources, including advanced ggplot2 techniques, check out these links:       

R Graphics ggplot2 Cookbook - <http://www.cookbook-r.com/Graphs/>     
Top 50 Visualizations - <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>    
R Studio's ggplot2 Cheat Sheet - <https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf>    
Hadley's ggplot2 book (amazon) - <https://www.amazon.com/dp/331924275X/ref=cm_sw_su_dp>    

<br  />

Thanks for reading,     
- Fisher 