---
title: "Unpacking the Tidyverse - tidyr"
author: "Fisher Ankney"
date: "February 23, 2018"
output: md_document
---
---
title: "Unpacking the Tidyverse - tidyr"
author: "Fisher Ankney"
date: "February 23, 2018"
output: html_document
---

<br> 

### Introduction 

This is the fourth of eight installments of my *Unpacking the Tidyverse* series. Each installment focuses on one of the eight core packages in Hadley Wickham's tidyverse. Instructions given in each post are mainly derived from Hadley's textbook, [R for Data Science](http://r4ds.had.co.nz/), and CRAN package documentation. This installment of *Unpacking the Tidyverse* focuses on the data-tidying package, tidyr. The previous installment focuses on the dplyr package, and can be found [here](link). The next installment focuses on the stringr package, and can be found [here](link). 

Tidy data is an easy and consistent way of storing data that makes future analytical steps simpler. Datasets that follow the three tidy data rules allow for R's vectorized nature to work its magic. The package tidyr is meant to coerce your dataset to follow the three rules of tidy data; it is not meant for general reshaping or aggregating of data. 

Tidy Data Rules: 

1. Each column is a variable <br  />
2. Each row is an observation <br  />
3. Each value has its own cell <br  />

<insert tidy data image here> 

<br> 

```{r, message=FALSE}
library('tidyverse')
```

<br> 

### Important Package Functions

<br> 

```
gather()        # Takes multiple columns and collapses into key-value pairs
spread()        # Spreads a key-value pair across multiple columns
separate()      # Single column into multiple columns, defined separator
unite()         # Multiple columns into single column, defined separator
complete()      # Turns implicit missing values explicit
fill()          # Fills in missing rows in a column based on the last entry
```

<br> 

### `gather()`

Often, a dataset has column names that are not the names of variables, but the values of variables. take `table4a` for example. 

```{r}
table4a
```

<br> 

In `table4a`, the column names `1999` and `2000` are values of the variable year, not variables themselves. The `gather()` function can fix this, it is perhaps the most commonly used function in tidyr. Begin by listing the column names that are actually variable names into `gather()` as its first arguments. Once these column names have been entered, define a `key` as the second argument. In this case, `1999` and `2000` are both years, so the key, or new variable name, is `year`. Finally, define a `value` to be associated with the key as the last argument in `gather()`. The `value` is the name given to the elements that were previously linked to the previous `1999` and `2000` false variables. Because these two years have been gathered into a new `year` variable, the associated elements need a new column to call home. In this case, the values represent a number positive cases, so the `value` is now defined as cases. 

```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

<br  />

### `spread()`


When an observation is scattered across multiple rows, variable names appear repeatedly as elements in a dataframe. In `table2` a single observation is a country in each year, but each observation is spread across two rows. 

```{r}
table2
```

<br  />

To tidy this data, utilize `spread()`, one of the most common functions in tidyr. The first argument in `spread()` is the column that contains the variable names, the `key` column. In the case of `table2`, the key column is type, because cases and population are both variables, not values. The second and final argument in `spread()` is the column that contains values associated with the previously defined key, the `value` column. In `table2` the it is the count column that contains the associated values.

```{r}
table2 %>%
spread(key = type, value = count)
```

<br  />

### `separate()`

Sometimes a column contains more than variable worth of information in each element. In `table3` this is exemplified by the `rate` column. 

```{r}
table3
```

<br  />

A simple problem to fix, the `separate()` function splits two variables apart from a single column. The first argument in `separate()` is the column in question, in this case it's `rate`. The next argument is a bit more complex; the `into` argument needs a concatenated input, `c()` to define the names and number of new columns. As for table3, the new columns are defined as `into = c("cases", "population")`. Finally, a separator character can be defined in the third argument. The separator defaults to any non-alphanumeric character, but can be customized. 

```{r, warning=FALSE}
table3 %>%
  separate(rate, into = c("cases", "population"))
```

<br  />

### `unite()`

When a single variable is spread between multiple columns, it's time to unite them. In `table5` the century and year variables should be united to create a single variable, the full year. 


```{r}
table5
```

<br  /> 

The `unite()` function works in a similar way to `separate()`, as it is it's inverse. The first argument of `unite()` is to define the name of the new, united, variable. In the case of `table3`, this variable shall be called `full_year`. The next arguments define the columns that need to be united; this can be any number of columns. `century` and `year` need to be united to form `full_year` in this example. Finally, it's sometimes wise to indicate a separator character, just as with `separate()`. The separator character in `unite()` defaults to the underscore, `_`. 

```{r}
table5 %>%
  unite(full_year, century, year, sep = "")
```

<br  /> 

### `complete()`

When dealing with missing data it's often important to turn implicitly missing values to explicit missing values. There are two missing values from the stocks tibble, 4th quarter 2015 and 1st quarter 2016. 

```{r, echo = FALSE}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
````

```{r}
stocks
```

<br> 

The `complete()` function takes a set of columns, and finds all unique combinations. It ensures the original dataset contains all those values, explicitly filling in `NA` when necessary. The input arguments of `complete()` are simply the columns you want to cross reference. In the case of `stocks` we want to find all of the combinations between the `year` and `qtr` variable, as to fill in implicit missing variables with `NA`. 

```{r}
stocks %>% 
  complete(year, qtr)
```

<br> 

### `fill()`

When dealing with missing data, it can be the case that you know that missing values are supposed to be carried on from the last observation. Something along the line of "ditto" quotations on a sign-up sheet. In the tibble treatment, we see just that. 

```{r, echo = FALSE}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
```

```{r}
treatment
```


The function `fill()` is the perfect fix for this situation. `fill()` takes a set of columns where you want missing values to be replaced with the most recent non-missing value. Simply input the column in question as the argument in `fill()`, and let R do the rest. In the case of the tibble `treatment`, the column in question is `person`. 

```{r}
treatment %>%
  fill(person)
```

<br> 

Additional Resources: 

- [CRAN Documentation](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
- [JSS: Tidy Data by Hadley Wickham](http://vita.had.co.nz/papers/tidy-data.pdf)
- [STHDA tidyr Tutorial](http://www.sthda.com/english/wiki/tidyr-crucial-step-reshaping-data-with-r-for-easier-analyses)

<br  />

Until next time, <br  />
\- Fisher


